---
title: "SDS Risk Manuscript Analysis Pipeline"
author: "Ryan Hamilton"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Quick start and run order

This annotated R Markdown file preserves the working manuscript analysis while making the pipeline easier to audit, rerun, and share on GitHub. The analysis is intentionally kept in one file so manuscript outputs can be traced from data import through models, figures, and supplemental tables.

**Expected repository layout**

```text
.
├── data/                 # Input data; not committed if restricted
├── figures/              # Exported manuscript figures
├── tables/               # Exported manuscript and supplemental tables
├── cache/                # Optional cached model objects
└── sds_risk_ms_pipeline_clean_annotated.Rmd
```

**Run order**

1. Setup paths, packages, helper functions, and output directories.
2. Import the 2020 to 2024 multistate SDS dataset and perform basic audits.
3. Generate Figure 1 from the separate mapping coordinate file.
4. Recode variables and build analysis-ready model datasets.
5. Evaluate assumptions and fit root disease, foliar disease, yield, and decision-window mixed models.
6. Generate correlation, association-network, PCA, and response-surface outputs.
7. Summarize ROI and decision-window results for Figure 5 and Supplemental Table S3.
8. Export manuscript figures and supplemental tables.
9. Record session information for reproducibility.

**Object naming guide**

- `df_raw`: imported and minimally filtered dataset.
- `df`: cleaned manuscript analysis dataset.
- `df_*`: analysis-specific datasets.
- `m_*`: fitted model objects.
- `emm_*`: estimated marginal means.
- `pairs_*`: pairwise contrast tables.
- `*_export`: final manuscript or supplemental table objects.

**Manuscript output map**

- Figure 1: trial-location map and site-year bubble summary.
- Figure 5: predicted yield response across pre-plant SDS risk levels.
- Table S1: cultivar deployment by state/province and year.
- Table S3: predicted yield and ROI across representative SDS risk levels.


```{r setup, include=FALSE}
# ============================================================
# Global knitr options and project paths
# ============================================================
# Cleaned, annotated version of the working manuscript pipeline.
# This file preserves the analytical workflow from the working Rmd and adds
# a GitHub-oriented structure for paths, outputs, and documentation.

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

options(scipen = 999)

# ---- User-editable paths ----
# Place input data in ./data/ for GitHub use, or edit locally.
data_path <- "data/full2020_2024.csv"
map_path  <- "data/map.csv"

# ---- Output directories ----
fig_dir   <- "figures"
table_dir <- "tables"
cache_dir <- "cache"

for (d in c(fig_dir, table_dir, cache_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

# ---- Lightweight helper functions ----
# These wrappers make export calls consistent and keep manuscript outputs in
# predictable folders. They are intentionally simple so the Rmd remains easy to
# run on a collaborator's machine.
export_plot <- function(plot, filename, width, height, dpi = 600, ...) {
  ggplot2::ggsave(
    filename = file.path(fig_dir, filename),
    plot = plot,
    width = width,
    height = height,
    dpi = dpi,
    ...
  )
}

export_table <- function(x, filename) {
  readr::write_csv(x, file.path(table_dir, filename))
}

standardize_product <- function(x) {
  dplyr::case_when(
    stringr::str_detect(as.character(x), stringr::regex("base|CB$", ignore_case = TRUE)) ~ "Base",
    stringr::str_detect(as.character(x), stringr::regex("ILEVO|fluopyram", ignore_case = TRUE)) ~ "ILEVO",
    stringr::str_detect(as.character(x), stringr::regex("Saltro|pydiflumetofen", ignore_case = TRUE)) ~ "Saltro",
    TRUE ~ as.character(x)
  )
}

standardize_sds_tol <- function(x) {
  dplyr::case_when(
    as.character(x) %in% c("r", "R") ~ "Moderately resistant",
    as.character(x) %in% c("s", "S") ~ "Susceptible",
    TRUE ~ as.character(x)
  )
}

getwd()
```

# ============================================================
# Multistate SDS risk manuscript pipeline
# Working dataset: full2020_2024.csv
# ============================================================
### Start clean
```{r clean-session, include=FALSE, results='hide', eval=FALSE}
# Optional interactive cleanup. This chunk is disabled by default for GitHub
# reproducibility because clearing the environment after setup removes path and
# helper objects defined above. Run manually only if needed.
rm(list = ls(all = TRUE))
```

# -----------------------------
# 0. Packages
# -----------------------------
```{r packages, warning=FALSE, include=FALSE}
# install_missing is FALSE by default for GitHub/reproducibility hygiene.
# Set to TRUE if running on a fresh local machine and you want auto-install.
install_missing <- FALSE
req_pkgs <- c(
  "tidyverse",
  "janitor",
  "readr",
  "stringr",
  "forcats",
  "lme4",
  "lmerTest",
  "emmeans",
  "multcomp",
  "multcompView",
  "performance",
  "DHARMa",
  "broom.mixed",
  "rstatix",
  "MASS",
  "car",
  "FactoMineR",
  "factoextra",
  "corrplot",
  "polycor",
  "DescTools",
  "igraph",
  "tidygraph",
  "ggraph",
  "scales",
  "patchwork",
  "mgcv"
)

# Packages used by the mapping and table-formatting sections.
req_pkgs <- unique(c(req_pkgs, "maps", "tidyr"))

to_install <- req_pkgs[!req_pkgs %in% installed.packages()[, "Package"]]
if (install_missing && length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}

invisible(lapply(req_pkgs, library, character.only = TRUE))

```

# -----------------------------
# 1. Import and basic audit
# -----------------------------
```{r include=FALSE}
df_raw <- read_csv(data_path, show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  filter(product %in% c("Base", "ILEVO", "Saltro")) %>%
  filter(!stringr::str_detect(location, "HEADSUP")) %>%
  mutate(
    location = stringr::str_remove(location, "_[0-9]+$")
  )

glimpse(df_raw)
dim(df_raw)
names(df_raw)

missing_tbl <- tibble(
  variable = names(df_raw),
  n_missing = sapply(df_raw, \(x) sum(is.na(x))),
  pct_missing = round(100 * n_missing / nrow(df_raw), 2)
) %>%
  arrange(desc(n_missing))

print(missing_tbl, n = nrow(missing_tbl))

state_counts <- df_raw %>% dplyr::count(state, sort = TRUE)
year_counts  <- df_raw %>% dplyr::count(year, sort = TRUE)

print(state_counts)
print(year_counts)


```

# -----------------------------
# 1A. Required-column validation
# -----------------------------
```{r required-column-validation}
# Fail early if a required manuscript variable is missing. This is especially
# helpful when collaborators rerun the pipeline with a revised flat file.
required_cols <- c(
  "state", "location", "year", "exp", "rep", "cultivar", "sds_tol",
  "product", "ratio_fvfg_dnang", "soil_ph", "scn_count_rep",
  "rdx", "fdx", "yield_kgha", "plant_date", "seed_rate_ha", "row_space_m"
)

missing_required <- setdiff(required_cols, names(df_raw))
if (length(missing_required) > 0) {
  stop(
    "Required columns are missing from df_raw: ",
    paste(missing_required, collapse = ", ")
  )
}

message("Required-column check passed: ", length(required_cols), " expected fields found.")
```

# -----------------------------
# 1B. Figure 1 mapping workflow
# -----------------------------
```{r figure-1-map, eval=FALSE}
# Purpose:
#   Generate the manuscript Figure 1 map showing field trial locations and the
#   number of site-years represented at each location. This block uses a separate
#   flat file (map.csv) containing latitude and longitude for each location.
#
# Required columns in map.csv:
#   state, location, year, lat, lon
#
# The chunk is eval=FALSE by default because map coordinates are maintained in a
# separate file and may not be distributed with restricted analysis data.

library(maps)
library(patchwork)

# ---- Read mapping file ----
df_map <- readr::read_csv(map_path, show_col_types = FALSE) %>%
  mutate(
    state = trimws(as.character(state)),
    location = trimws(as.character(location)),
    region_panel = case_when(
      state == "ON" ~ "Ontario",
      state == "DE" ~ "Delaware",
      state %in% c("MI", "IL", "IN", "OH", "WI", "IA", "MO", "KS", "KY", "ND", "NE") ~ "North Central U.S.",
      TRUE ~ "Other"
    )
  ) %>%
  filter(region_panel != "Other")

# ---- Count site-years per unique location ----
# Site-years are defined as unique location × year combinations. This is the
# quantity encoded by bubble size in Figure 1.
df_counts <- df_map %>%
  group_by(region_panel, state, location) %>%
  summarise(
    site_years = n_distinct(year),
    lat = first(lat),
    lon = first(lon),
    .groups = "drop"
  )

# ---- Basemap data ----
us_map <- map_data("state")
can_map <- map_data("world") %>%
  filter(region == "Canada")

# ---- Shared map style ----
map_theme <- theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    plot.margin = margin(t = 0, r = 2, b = 0, l = 4),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey85", linewidth = 0.2),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 9),
    axis.ticks = element_line(linewidth = 0.2),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6)
  )

siteyear_limits <- range(df_counts$site_years, na.rm = TRUE)
size_breaks <- sort(unique(df_counts$site_years))

# ---- North Central U.S. panel ----
p_midwest <- ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92",
    color = "grey40",
    linewidth = 0.25
  ) +
  geom_point(
    data = df_counts %>% filter(region_panel == "North Central U.S."),
    aes(x = lon, y = lat, size = site_years),
    shape = 21,
    fill = "grey60",
    color = "black",
    alpha = 0.7,
    stroke = 0.3
  ) +
  coord_quickmap(xlim = c(-100, -80.1), ylim = c(36, 49)) +
  scale_size_continuous(
    range = c(2.5, 8),
    limits = siteyear_limits,
    breaks = size_breaks,
    name = "Site-years",
    guide = "legend"
  ) +
  scale_x_continuous(expand = expansion(mult = 0)) +
  scale_y_continuous(expand = expansion(mult = 0)) +
  labs(title = "North Central U.S.A.", x = "Longitude", y = "Latitude") +
  map_theme +
  theme(plot.title = element_text(face = "bold", hjust = 0, vjust = -7, margin = margin(b = 0)))

# ---- Ontario panel ----
p_ontario <- ggplot() +
  geom_polygon(
    data = can_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92",
    color = "grey40",
    linewidth = 0.25
  ) +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92",
    color = "grey40",
    linewidth = 0.25
  ) +
  geom_point(
    data = df_counts %>% filter(region_panel == "Ontario"),
    aes(x = lon, y = lat, size = site_years),
    shape = 21,
    fill = "grey60",
    color = "black",
    alpha = 0.7,
    stroke = 0.3
  ) +
  coord_quickmap(xlim = c(-84.75, -80), ylim = c(40.75, 44.25)) +
  scale_size_continuous(
    range = c(2.5, 8),
    limits = siteyear_limits,
    breaks = size_breaks,
    name = "Site-years",
    guide = "none"
  ) +
  scale_x_continuous(expand = expansion(mult = 0)) +
  scale_y_continuous(expand = expansion(mult = 0)) +
  labs(title = "Ontario, Canada", x = "Longitude", y = "Latitude") +
  map_theme +
  theme(plot.title = element_text(face = "bold", hjust = 0, vjust = -8, margin = margin(l = 4)))

# ---- Delaware panel ----
p_delaware <- ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "grey92",
    color = "grey40",
    linewidth = 0.25
  ) +
  geom_point(
    data = df_counts %>% filter(region_panel == "Delaware"),
    aes(x = lon, y = lat, size = site_years),
    shape = 21,
    fill = "grey60",
    color = "black",
    alpha = 0.7,
    stroke = 0.3
  ) +
  coord_quickmap(xlim = c(-76.5, -75), ylim = c(38.25, 40)) +
  scale_size_continuous(
    range = c(2.5, 8),
    limits = siteyear_limits,
    breaks = size_breaks,
    name = "Site-years",
    guide = "none"
  ) +
  scale_x_continuous(expand = expansion(mult = 0)) +
  scale_y_continuous(expand = expansion(mult = 0)) +
  labs(title = "Delaware, U.S.A.", x = "Longitude", y = "Latitude") +
  map_theme +
  theme(plot.title = element_text(face = "bold", hjust = 0, vjust = -6, margin = margin(l = 6)))

# ---- Assemble panels ----
right_panels <- p_ontario / patchwork::plot_spacer() / p_delaware +
  plot_layout(heights = c(1, 0.03, 1.25))

p_map <- (p_midwest | patchwork::plot_spacer() | right_panels) +
  plot_layout(widths = c(2.5, 0.12, 0.75), guides = "collect") &
  theme(
    legend.position = "right",
    legend.box = "vertical",
    legend.justification = "center",
    plot.margin = margin(0, 2.5, 0, 2.5)
  )

p_map

ggsave(
  file.path(fig_dir, "Figure_1_trial_locations.png"),
  p_map,
  width = 9,
  height = 6,
  dpi = 600
)
```
# -----------------------------
# 2. Preprocess / recode
# -----------------------------
```{r}
df <- df_raw %>%
  mutate(
    # Factors
    state    = as.factor(state),
    location = as.factor(location),
    year     = as.factor(year),
    exp      = as.factor(exp),
    rep      = as.factor(rep),
    cultivar = as.factor(cultivar),
    product  = as.factor(product),
    sds_tol  = as.factor(sds_tol),

    # Hierarchical helpers
    siteyear = factor(paste(location, year, sep = "_")),
    
    # Numeric coercion
    ratio_fvfg_dnang = as.numeric(ratio_fvfg_dnang),
    scn_count_rep    = as.numeric(scn_count_rep),
    soil_p_h         = as.numeric(soil_ph),
    fdx              = as.numeric(fdx),
    rdx              = as.numeric(rdx),
    drw              = as.numeric(drw),
    pl_pop_ha        = as.numeric(pl_pop_ha),
    yield_kgha       = as.numeric(yield_kgha),
    proj_price_kg    = as.numeric(proj_price_kg),
    harv_price_kg    = as.numeric(harv_price_kg),
    volatility       = as.numeric(volatility),

    # Transformations for LMM fitting
    log10_fv   = log10(ratio_fvfg_dnang + 0.001),
    log10_scn  = log10(scn_count_rep + 1),
    log1p_fdx  = log1p(fdx),
    log1p_rdx  = log1p(rdx),
    log1p_drw  = log1p(drw),

    # Disease category
    fdx_level = case_when(
      is.na(fdx)            ~ NA_character_,
      fdx == 0              ~ "none",
      fdx > 0 & fdx < 10    ~ "low",
      fdx >= 10             ~ "high"
    ) %>% factor(levels = c("none", "low", "high")),

    # SCN category
    scn_level = case_when(
      is.na(scn_count_rep)  ~ NA_character_,
      scn_count_rep == 0    ~ "none",
      scn_count_rep < 2000  ~ "low",
      scn_count_rep >= 2000 ~ "high"
    ) %>% factor(levels = c("none", "low", "high")),

    # Condensed neutral Fv bins for plotting / trees
    fv_level_simple = case_when(
      is.na(ratio_fvfg_dnang)                     ~ NA_character_,
      ratio_fvfg_dnang == 0                       ~ "0",
      ratio_fvfg_dnang > 0 & ratio_fvfg_dnang < 1 ~ "0-1",
      ratio_fvfg_dnang >= 1 & ratio_fvfg_dnang < 5  ~ "1-4",
      ratio_fvfg_dnang >= 5                         ~ ">5"
    ) %>% factor(levels = c("0", "0-1", "1-4", ">5"))
  )

summary(df %>% select(ratio_fvfg_dnang, scn_count_rep, soil_ph, fdx, yield_kgha))
table(df$fdx_level, useNA = "ifany")
table(df$scn_level, useNA = "ifany")
table(df$fv_level_simple, useNA = "ifany")

```
# -----------------------------
# 3. Exploratory plots
# -----------------------------
```{r warning=FALSE}
p_hist_fv <- ggplot(df, aes(x = ratio_fvfg_dnang)) +
  geom_histogram(bins = 40) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "F. virguliforme DNA abundance", x = "fg Fv DNA / ng total eDNA", y = "Count")

p_hist_scn <- ggplot(df, aes(x = scn_count_rep)) +
  geom_histogram(bins = 40) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "SCN abundance", x = "SCN eggs + J2 / 100 cm^3", y = "Count")

p_hist_fdx <- ggplot(df, aes(x = fdx)) +
  geom_histogram(bins = 40) +
  labs(title = "FDX distribution", x = "FDX", y = "Count")

p_hist_yield <- ggplot(df, aes(x = yield_kgha)) +
  geom_histogram(bins = 40) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = "Yield distribution", x = "Yield (kg/ha)", y = "Count")

(p_hist_fv | p_hist_scn) / (p_hist_fdx | p_hist_yield)

ggsave("exploratory_histograms.png", width = 12, height = 8, dpi = 300)

```

# -----------------------------
# 4. Core analysis datasets
# -----------------------------
```{r}
df_fdx <- df %>%
  filter(
    !is.na(fdx),
    !is.na(log10_fv),
    !is.na(log10_scn),
    !is.na(product),
    !is.na(sds_tol)
  )

df_yield <- df %>%
  filter(
    !is.na(yield_kgha),
    !is.na(log10_fv),
    !is.na(log10_scn),
    !is.na(product),
    !is.na(sds_tol)
  )

df_fdx_ph <- df_fdx %>% filter(!is.na(soil_ph))
df_yield_ph <- df_yield %>% filter(!is.na(soil_ph))

```

### Sanity check ###
```{r}
glimpse(df)

summary(df %>%
  select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ))

df %>% 
  dplyr::count(siteyear) %>% 
  arrange(desc(n))

df %>%
  dplyr::count(siteyear, product) %>%
  pivot_wider(names_from = product, values_from = n)
```
# -----------------------------
# 4B. Root subset datasets and diagnostics
# -----------------------------
```{r}
df_root <- df %>%
  filter(
    !is.na(rdx),
    !is.na(drw),
    !is.na(log10_fv),
    !is.na(log10_scn),
    !is.na(product),
    !is.na(sds_tol)
  )

df_root_ph <- df_root %>%
  filter(!is.na(soil_ph))

cat("\n================ Root subset overview ================\n")
print(
  df_root %>%
    summarise(
      n_obs = n(),
      n_siteyear = n_distinct(siteyear),
      n_location = n_distinct(location),
      n_year = n_distinct(year)
    )
)

cat("\nObservations by product in root subset:\n")
print(df_root %>% dplyr::count(product, sort = TRUE))

cat("\nObservations by SDS tolerance in root subset:\n")
print(df_root %>% dplyr::count(sds_tol, sort = TRUE))

cat("\nObservations by siteyear in root subset:\n")
print(df_root %>% dplyr::count(siteyear, sort = TRUE))

# Quick summaries
cat("\n================ Root subset summaries ================\n")
print(summary(df_root %>% select(rdx, drw, ratio_fvfg_dnang, scn_count_rep, fdx, yield_kgha)))

# Histograms
p_hist_rdx <- ggplot(df_root, aes(x = rdx)) +
  geom_histogram(bins = 40) +
  labs(title = "RDX distribution", x = "Root disease index (RDX)", y = "Count")

p_hist_drw <- ggplot(df_root, aes(x = drw)) +
  geom_histogram(bins = 40) +
  labs(title = "DRW distribution", x = "Dry root weight (DRW)", y = "Count")

p_hist_rdx_log <- ggplot(df_root, aes(x = log1p_rdx)) +
  geom_histogram(bins = 40) +
  labs(title = "log1p(RDX) distribution", x = "log1p(RDX)", y = "Count")

p_hist_drw_log <- ggplot(df_root, aes(x = log1p_drw)) +
  geom_histogram(bins = 40) +
  labs(title = "log1p(DRW) distribution", x = "log1p(DRW)", y = "Count")

(p_hist_rdx | p_hist_drw) / (p_hist_rdx_log | p_hist_drw_log)

ggsave("exploratory_root_histograms.png", width = 12, height = 8, dpi = 300)

# -----------------------------
# Root subset assumption testing
# -----------------------------
cat("\n================ Root subset Levene's tests ================\n")
print(car::leveneTest(rdx ~ product, data = df_root))
print(car::leveneTest(drw ~ product, data = df_root))
print(car::leveneTest(rdx ~ sds_tol, data = df_root))
print(car::leveneTest(drw ~ sds_tol, data = df_root))

# Reuse safe_shapiro if already defined above; if not, define here
safe_shapiro <- function(x, label = "variable") {
  x <- x[!is.na(x)]
  if (length(x) > 5000) {
    set.seed(123)
    x <- sample(x, 5000)
  }
  out <- shapiro.test(x)
  cat("\nShapiro-Wilk test for", label, "\n")
  print(out)
}

safe_shapiro(df_root$rdx, "rdx (raw)")
safe_shapiro(df_root$log1p_rdx, "rdx (log1p)")
safe_shapiro(df_root$drw, "drw (raw)")
safe_shapiro(df_root$log1p_drw, "drw (log1p)")

# Box-Cox on positive values only
cat("\n================ Root subset Box-Cox tests ================\n")

df_rdx_pos <- df_root %>% filter(rdx > 0)
if (nrow(df_rdx_pos) > 20) {
  bc_rdx <- MASS::boxcox(
    lm(rdx ~ log10_fv + log10_scn + product + sds_tol, data = df_rdx_pos),
    plotit = FALSE
  )
  lambda_rdx <- bc_rdx$x[which.max(bc_rdx$y)]
  cat("Estimated Box-Cox lambda for positive RDX:", lambda_rdx, "\n")
} else {
  lambda_rdx <- NA_real_
  cat("Not enough positive RDX values for Box-Cox estimation.\n")
}

df_drw_pos <- df_root %>% filter(drw > 0)
if (nrow(df_drw_pos) > 20) {
  bc_drw <- MASS::boxcox(
    lm(drw ~ log10_fv + log10_scn + product + sds_tol, data = df_drw_pos),
    plotit = FALSE
  )
  lambda_drw <- bc_drw$x[which.max(bc_drw$y)]
  cat("Estimated Box-Cox lambda for positive DRW:", lambda_drw, "\n")
} else {
  lambda_drw <- NA_real_
  cat("Not enough positive DRW values for Box-Cox estimation.\n")
}

# Simple biological relationships for later mechanistic interpretation
cat("\n================ Root subset simple correlations ================\n")
print(cor.test(df_root$rdx, df_root$fdx, method = "spearman", use = "pairwise.complete.obs"))
print(cor.test(df_root$rdx, df_root$yield_kgha, method = "spearman", use = "pairwise.complete.obs"))
print(cor.test(df_root$drw, df_root$fdx, method = "spearman", use = "pairwise.complete.obs"))
print(cor.test(df_root$drw, df_root$yield_kgha, method = "spearman", use = "pairwise.complete.obs"))

# Quick exploratory plots
p_rdx_fdx <- ggplot(df_root, aes(x = rdx, y = fdx)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "RDX vs FDX", x = "RDX", y = "FDX")

p_rdx_yield <- ggplot(df_root, aes(x = rdx, y = yield_kgha)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "RDX vs yield", x = "RDX", y = "Yield (kg/ha)")

p_drw_fdx <- ggplot(df_root, aes(x = drw, y = fdx)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "DRW vs FDX", x = "DRW", y = "FDX")

p_drw_yield <- ggplot(df_root, aes(x = drw, y = yield_kgha)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "DRW vs yield", x = "DRW", y = "Yield (kg/ha)")

(p_rdx_fdx | p_rdx_yield) / (p_drw_fdx | p_drw_yield)

ggsave("exploratory_root_relationships.png", width = 12, height = 8, dpi = 300)

df_root %>% dplyr::count(siteyear)
write_csv(df, "full2020_2024_analysis_ready.csv")

```

# -----------------------------
# 5. Formal assumption testing
# -----------------------------
```{r}
# Levene's tests (homogeneity of variance)
# These are most interpretable for grouped factors such as product and tolerance
cat("\n================ Levene's tests ================\n")
print(car::leveneTest(fdx ~ product, data = df_fdx))
print(car::leveneTest(yield_kgha ~ product, data = df_yield))
print(car::leveneTest(fdx ~ sds_tol, data = df_fdx))
print(car::leveneTest(yield_kgha ~ sds_tol, data = df_yield))

# Shapiro-Wilk tests
# Run on raw and transformed responses; use a sample if n > 5000
safe_shapiro <- function(x, label = "variable") {
  x <- x[!is.na(x)]
  if (length(x) > 5000) {
    set.seed(123)
    x <- sample(x, 5000)
  }
  out <- shapiro.test(x)
  cat("\nShapiro-Wilk test for", label, "\n")
  print(out)
}

safe_shapiro(df_fdx$fdx, "fdx (raw)")
safe_shapiro(df_fdx$log1p_fdx, "fdx (log1p)")
safe_shapiro(df_yield$yield_kgha, "yield_kgha (raw)")
#safe_shapiro(df_yield$log_yield, "yield_kgha (log)")
safe_shapiro(df_yield$yield_kgha, "yield_kgha (for Box-Cox reference)")

# Box-Cox tests for yield
# Box-Cox requires strictly positive response
cat("\n================ Box-Cox tests ================\n")
#bc_yield <- MASS::boxcox(
  #lm(yield_kgha ~ log10_fv + log10_scn + product + sds_tol, data = df_yield),
  #plotit = FALSE
#)
#lambda_yield <- bc_yield$x[which.max(bc_yield$y)]
#cat("Estimated Box-Cox lambda for yield_kgha:", lambda_yield, "\n")

# Apply Box-Cox transformed yield for modeling
#if (abs(lambda_yield) < 1e-6) {
 #df_yield <- df_yield %>% mutate(yield_bc = log(yield_kgha))
#} else {
 # df_yield <- df_yield %>%
    #mutate(yield_bc = (yield_kgha^lambda_yield - 1) / lambda_yield)
#}

# Optional Box-Cox for positive disease subsets only
df_fdx_pos <- df_fdx %>% filter(fdx > 0)
if (nrow(df_fdx_pos) > 20) {
  bc_fdx <- MASS::boxcox(
    lm(fdx ~ log10_fv + log10_scn + product + sds_tol, data = df_fdx_pos),
    plotit = FALSE
  )
  lambda_fdx <- bc_fdx$x[which.max(bc_fdx$y)]
  cat("Estimated Box-Cox lambda for positive FDX:", lambda_fdx, "\n")
} else {
  lambda_fdx <- NA_real_
  cat("Not enough positive FDX values for Box-Cox estimation.\n")
}

```

# -----------------------------
# 6. Linear mixed models
# -----------------------------
# Core model strategy:
# - Use transformed response variables where diagnostics justify it
# - Use Tukey HSD only for pairwise comparisons
# - No BH / FDR adjustments in the inferential core

# -------------------------------
# 6a. RDX models - df_root subset
# -------------------------------
```{r}

df_root_mod <- df_root %>%
  mutate(
    z_log10_fv   = as.numeric(scale(log10_fv)),
    z_log10_scn  = as.numeric(scale(log10_scn)),
    exp_siteyear = interaction(siteyear, exp, drop = TRUE)
  )

df_root_ph_mod <- df_root_ph %>%
  mutate(
    z_log10_fv   = as.numeric(scale(log10_fv)),
    z_log10_scn  = as.numeric(scale(log10_scn)),
    z_soil_ph    = as.numeric(scale(soil_ph)),
    exp_siteyear = interaction(siteyear, exp, drop = TRUE)
  )

conflicts_prefer(lme4::lmer)

# -----------------------------
# RDX models
# -----------------------------

m_rdx_1 <- lmer(
  log1p_rdx ~ z_log10_fv + z_log10_scn + sds_tol +
    (1 | siteyear),
  data = df_root_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

cat("\n================ RDX model 1 ================\n")
print(summary(m_rdx_1))
print(emmeans::joint_tests(m_rdx_1))
print(performance::check_singularity(m_rdx_1))
print(performance::check_collinearity(m_rdx_1))

# Optional pH model
m_rdx_2 <- lmer(
  log1p_rdx ~ z_log10_fv + z_log10_scn + z_soil_ph + sds_tol:product + (1 | siteyear),
  data = df_root_ph_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

cat("\n================ RDX model 2 (with pH) ================\n")
print(summary(m_rdx_2))
print(emmeans::joint_tests(m_rdx_2))
print(performance::check_singularity(m_rdx_2))
print(performance::check_collinearity(m_rdx_2))

# EMMs for product from core model
emm_rdx_product <- emmeans(m_rdx_2, ~ product, type = "response")
multcomp::cld(emm_rdx_product)

rdx_emm_bt <- as.data.frame(emm_rdx_product) %>%
  dplyr::mutate(
    emmean_bt   = exp(emmean) - 1,
    lower.CL_bt = exp(lower.CL) - 1,
    upper.CL_bt = exp(upper.CL) - 1
  )

rdx_emm_bt_fmt <- rdx_emm_bt %>%
  dplyr::transmute(
    product,
    EMM_bt = round(emmean_bt, 2),
    lower.CL_bt = round(lower.CL_bt, 2),
    upper.CL_bt = round(upper.CL_bt, 2)
  )


# Refit no-pH model on same rows for valid AIC comparison
m_rdx_1_same_rows <- lmer(
  log1p_rdx ~ z_log10_fv +
    (1 | siteyear),
  data = df_root_ph_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

cat("\n================ RDX AIC comparison on same rows ================\n")
print(AIC(m_rdx_1_same_rows, m_rdx_2))

# EMMs for product
#emm_rdx_product <- emmeans(m_rdx_1, ~ product)

#cat("\n================ RDX product EMMs ================\n")
#print(emm_rdx_product)
#print(pairs(emm_rdx_product, adjust = "tukey"))
#print(cld(emm_rdx_product, Letters = letters, adjust = "tukey"))

# Optional EMMs for sds_tol
#emm_rdx_tol <- emmeans(m_rdx_1, ~ sds_tol)

#cat("\n================ RDX sds_tol EMMs ================\n")
#print(emm_rdx_tol)
#print(pairs(emm_rdx_tol, adjust = "tukey"))
#print(cld(emm_rdx_tol, Letters = letters, adjust = "tukey"))

# -----------------------------
# DRW models
# -----------------------------

#m_drw_1 <- lmer(
  #log1p_drw ~ z_log10_fv + z_log10_scn + product + sds_tol +
    #(1 | siteyear) + (1 | exp_siteyear),
  # = df_root_mod,
  # = FALSE,
  # = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
#)

#cat("\n================ DRW model 1 ================\n")
#print(summary(m_drw_1))
#print(emmeans::joint_tests(m_drw_1))
#print(performance::check_singularity(m_drw_1))
#print(performance::check_collinearity(m_drw_1))

# Optional pH model
#m_drw_2 <- lmer(
  #log1p_drw ~ z_log10_fv + z_log10_scn + z_soil_ph + product + sds_tol +
    #(1 | siteyear) + (1 | exp_siteyear),
  #data = df_root_ph_mod,
  #REML = FALSE,
  #control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
#)

#cat("\n================ DRW model 2 (with pH) ================\n")
#print(summary(m_drw_2))
#print(emmeans::joint_tests(m_drw_2))
#print(performance::check_singularity(m_drw_2))
#print(performance::check_collinearity(m_drw_2))

# Refit no-pH model on same rows for valid AIC comparison
#m_drw_1_same_rows <- lmer(
  #log1p_drw ~ z_log10_fv + z_log10_scn + product + sds_tol +
    #(1 | siteyear) + (1 | exp_siteyear),
  #data = df_root_ph_mod,
  #REML = FALSE,
 ## control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
#)

#cat("\n================ DRW AIC comparison on same rows ================\n")
#print(AIC(m_drw_1_same_rows, m_drw_2))

# EMMs for product
#emm_drw_product <- emmeans(m_drw_1, ~ product)

#cat("\n================ DRW product EMMs ================\n")
#print(emm_drw_product)
#print(pairs(emm_drw_product, adjust = "tukey"))
#print(cld(emm_drw_product, Letters = letters, adjust = "tukey"))

# Optional EMMs for sds_tol
#emm_drw_tol <- emmeans(m_drw_1, ~ sds_tol)

#cat("\n================ DRW sds_tol EMMs ================\n")
#print(emm_drw_tol)
#print(pairs(emm_drw_tol, adjust = "tukey"))
#print(cld(emm_drw_tol, Letters = letters, adjust = "tukey"))
```


#-----------------------------
# Theme (APS-style, consistent across figures)
#-----------------------------
```{r}

theme_siteyear <- ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(linewidth = 0.2),
    #plot.title = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(color = "black"),
    axis.text = ggplot2::element_text(color = "black"),
    strip.text = ggplot2::element_text(color = "black"),
    #legend.title = ggplot2::element_blank(),
    legend.position = "right"
  )
```



# ============================================================
# 6b. Root subset downstream analyses -- 
# Correlation, association matrix, mini-network, PCA
# ============================================================
# -----------------------------
# 1. Root subset numeric correlations
# -----------------------------
```{r}
root_num <- df_root %>%
  select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  )

# Spearman correlation matrix
root_cor <- cor(root_num, method = "spearman", use = "pairwise.complete.obs")
print(root_cor)

# Pairwise Spearman tests
root_cor_tests <- expand.grid(
  var1 = names(root_num),
  var2 = names(root_num),
  stringsAsFactors = FALSE
) %>%
  filter(var1 < var2) %>%
  rowwise() %>%
  mutate(
    test = list(
      suppressWarnings(
        cor.test(
          root_num[[var1]],
          root_num[[var2]],
          method = "spearman",
          use = "pairwise.complete.obs"
        )
      )
    ),
    rho = unname(test$estimate),
    p_value = test$p.value
  ) %>%
  ungroup() %>%
  select(var1, var2, rho, p_value)

cat("\n================ Root subset pairwise Spearman correlations ================\n")
print(root_cor_tests)

write_csv(root_cor_tests, "root_pairwise_spearman_correlations.csv")

# Visualize numeric correlation matrix
png("fig_root_numeric_correlation_matrix.png", width = 1600, height = 1400, res = 220)
corrplot::corrplot(
  root_cor,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.7,
  mar = c(0, 0, 2, 0),
  title = "Root subset numeric Spearman correlation matrix"
)
dev.off()

```
# -----------------------------
# 2. Mixed-type association matrix for root subset
# -----------------------------
```{r}
# Helper: normalize all result objects to the same shape
safe_extract_result <- function(res) {
  out <- c(estimate = NA_real_, p = NA_real_)

  # NULL / empty
  if (is.null(res) || length(res) == 0) {
    return(out)
  }

  # Bare numeric scalar
  if (is.numeric(res) && length(res) == 1 && is.null(names(res))) {
    out["estimate"] <- as.numeric(res)
    return(out)
  }

  # Named atomic vector
  if (is.atomic(res) && !is.null(names(res))) {
    if ("estimate" %in% names(res) && length(res[["estimate"]]) == 1) {
      out["estimate"] <- suppressWarnings(as.numeric(res[["estimate"]]))
    }
    if ("p" %in% names(res) && length(res[["p"]]) == 1) {
      out["p"] <- suppressWarnings(as.numeric(res[["p"]]))
    }
    return(out)
  }

  # List with named elements
  if (is.list(res)) {
    if (!is.null(res$estimate) && length(res$estimate) == 1) {
      out["estimate"] <- suppressWarnings(as.numeric(res$estimate))
    }
    if (!is.null(res$p) && length(res$p) == 1) {
      out["p"] <- suppressWarnings(as.numeric(res$p))
    }
    return(out)
  }

  out
}

safe_cor_test <- function(x, y, method = "pearson") {
  ok <- stats::complete.cases(x, y)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  out <- tryCatch(
    suppressWarnings(stats::cor.test(x[ok], y[ok], method = method)),
    error = function(e) NULL
  )

  if (is.null(out)) return(c(estimate = NA_real_, p = NA_real_))

  c(
    estimate = unname(out$estimate),
    p = out$p.value
  )
}

safe_eta <- function(num, fac) {
  ok <- stats::complete.cases(num, fac)
  if (sum(ok) < 5 || nlevels(factor(fac[ok])) < 2) {
    return(c(estimate = NA_real_, p = NA_real_))
  }

  fit <- tryCatch(
    stats::aov(num[ok] ~ factor(fac[ok])),
    error = function(e) NULL
  )

  if (is.null(fit)) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    DescTools::EtaSq(fit, type = 1)[1, "eta.sq"],
    error = function(e) NA_real_
  )

  p <- tryCatch(
    summary(fit)[[1]][["Pr(>F)"]][1],
    error = function(e) NA_real_
  )

  c(estimate = est, p = p)
}

safe_cramers_v <- function(x, y) {
  ok <- stats::complete.cases(x, y)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  tab <- table(x[ok], y[ok])
  if (min(dim(tab)) < 2) return(c(estimate = NA_real_, p = NA_real_))

  chi <- tryCatch(
    suppressWarnings(stats::chisq.test(tab)),
    error = function(e) NULL
  )

  est <- tryCatch(
    DescTools::CramerV(tab),
    error = function(e) NA_real_
  )

  p <- if (is.null(chi)) NA_real_ else chi$p.value

  c(estimate = est, p = p)
}

safe_polyserial <- function(x, y_ord) {
  ok <- stats::complete.cases(x, y_ord)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    polycor::polyserial(x[ok], as.ordered(y_ord[ok]), std.err = FALSE),
    error = function(e) NA_real_
  )

  c(estimate = est, p = NA_real_)
}

safe_polychor <- function(x_ord, y_ord) {
  ok <- stats::complete.cases(x_ord, y_ord)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    polycor::polychor(x_ord[ok], y_ord[ok]),
    error = function(e) NA_real_
  )

  c(estimate = est, p = NA_real_)
}

root_assoc_df <- df_root %>%
  dplyr::transmute(
    ratio_fvfg_dnang = ratio_fvfg_dnang,
    scn_count_rep    = scn_count_rep,
    soil_ph          = soil_ph,
    rdx              = rdx,
    drw              = drw,
    fdx              = fdx,
    yield_kgha       = yield_kgha,
    product          = as.factor(product),
    sds_tol          = as.factor(sds_tol),
    scn_level        = factor(scn_level, ordered = TRUE),
    fv_level_simple  = factor(fv_level_simple, ordered = TRUE)
  )

root_var_names <- names(root_assoc_df)

root_var_type <- c(
  ratio_fvfg_dnang = "numeric",
  scn_count_rep    = "numeric",
  soil_ph          = "numeric",
  rdx              = "numeric",
  drw              = "numeric",
  fdx              = "numeric",
  yield_kgha       = "numeric",
  product          = "nominal",
  sds_tol          = "nominal",
  scn_level        = "ordinal",
  fv_level_simple  = "ordinal"
)

root_assoc_mat <- matrix(
  NA_real_,
  nrow = length(root_var_names),
  ncol = length(root_var_names),
  dimnames = list(root_var_names, root_var_names)
)

root_p_mat <- matrix(
  NA_real_,
  nrow = length(root_var_names),
  ncol = length(root_var_names),
  dimnames = list(root_var_names, root_var_names)
)

for (i in seq_along(root_var_names)) {
  for (j in seq_along(root_var_names)) {

    xi <- root_assoc_df[[root_var_names[i]]]
    xj <- root_assoc_df[[root_var_names[j]]]
    ti <- root_var_type[[root_var_names[i]]]
    tj <- root_var_type[[root_var_names[j]]]

    if (i == j) {
      root_assoc_mat[i, j] <- 1
      root_p_mat[i, j] <- 0
      next
    }

    res <- c(estimate = NA_real_, p = NA_real_)

    if (ti == "numeric" && tj == "numeric") {
      res <- safe_cor_test(xi, xj, method = "pearson")

    } else if (ti == "numeric" && tj == "nominal") {
      res <- safe_eta(xi, xj)

    } else if (ti == "nominal" && tj == "numeric") {
      res <- safe_eta(xj, xi)

    } else if (ti == "nominal" && tj == "nominal") {
      res <- safe_cramers_v(xi, xj)

    } else if (ti == "numeric" && tj == "ordinal") {
      res <- safe_polyserial(xi, xj)

    } else if (ti == "ordinal" && tj == "numeric") {
      res <- safe_polyserial(xj, xi)

    } else if (ti == "ordinal" && tj == "ordinal") {
      res <- safe_polychor(xi, xj)

    } else if (ti == "nominal" && tj == "ordinal") {
      res <- safe_cramers_v(xi, xj)

    } else if (ti == "ordinal" && tj == "nominal") {
      res <- safe_cramers_v(xi, xj)
    }

    res_clean <- safe_extract_result(res)

    # force scalar numeric values
    est_val <- if (length(res_clean["estimate"]) == 1) as.numeric(res_clean["estimate"]) else NA_real_
    p_val   <- if (length(res_clean["p"]) == 1) as.numeric(res_clean["p"]) else NA_real_

    root_assoc_mat[i, j] <- est_val
    root_p_mat[i, j] <- p_val
  }
}

readr::write_csv(
  as.data.frame(root_assoc_mat) %>% tibble::rownames_to_column("variable"),
  "root_association_matrix.csv"
)

readr::write_csv(
  as.data.frame(root_p_mat) %>% tibble::rownames_to_column("variable"),
  "root_association_pvalues.csv"
)

png("fig_root_association_matrix.png", width = 1700, height = 1500, res = 220)
corrplot::corrplot(
  root_assoc_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  na.label = " ",
  mar = c(0, 0, 2, 0),
  title = "Root subset mixed-type association matrix"
)
dev.off()
```
# -----------------------------
# 3. Mini-network association mapping
# -----------------------------
```{r}
# Numeric-focused "mini-network" for the root pathway
root_net_df <- df_root %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

root_net_cor <- stats::cor(
  root_net_df,
  method = "spearman",
  use = "pairwise.complete.obs"
)

root_net_p <- matrix(
  NA_real_,
  nrow = ncol(root_net_df),
  ncol = ncol(root_net_df),
  dimnames = list(names(root_net_df), names(root_net_df))
)

for (i in seq_len(ncol(root_net_df))) {
  for (j in seq_len(ncol(root_net_df))) {
    if (i == j) {
      root_net_p[i, j] <- 0
    } else {
      ct <- suppressWarnings(
        stats::cor.test(
          root_net_df[[i]],
          root_net_df[[j]],
          method = "spearman"
        )
      )
      root_net_p[i, j] <- ct$p.value
    }
  }
}

root_edge_tbl <- as.data.frame(as.table(root_net_cor)) %>%
  dplyr::rename(from = Var1, to = Var2, weight = Freq) %>%
  dplyr::mutate(p = as.vector(root_net_p)) %>%
  dplyr::filter(from != to) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(pair = paste(sort(c(from, to)), collapse = "__")) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(pair, .keep_all = TRUE) %>%
  dplyr::filter(!is.na(weight), abs(weight) >= 0.10, p < 0.05) %>%
  dplyr::mutate(
    layout_weight = abs(weight),
    sign_class = factor(ifelse(weight > 0, "positive", "negative"),
                        levels = c("positive", "negative"))
  )

root_node_tbl <- tibble::tibble(
  name = colnames(root_net_cor),
  category = factor(
    c("pathogen", "pathogen", "soil", "disease", "root", "disease", "yield"),
    levels = c("pathogen", "soil", "root", "disease", "yield")
  )
)

root_graph_obj <- tidygraph::tbl_graph(
  nodes = root_node_tbl,
  edges = root_edge_tbl,
  directed = FALSE
)

p_root_net <- ggraph::ggraph(
  root_graph_obj,
  layout = "fr",
  weights = layout_weight
) +
  ggraph::geom_edge_link(
    ggplot2::aes(
      edge_width = abs(weight),
      edge_alpha = abs(weight),
      edge_colour = sign_class,
      edge_linetype = sign_class
    )
  ) +
  ggraph::geom_node_point(
    ggplot2::aes(shape = category),
    size = 8
  ) +
  ggraph::geom_node_text(
    ggplot2::aes(label = name),
    repel = TRUE
  ) +
  ggraph::scale_edge_width(range = c(0.5, 2.5)) +
  ggraph::scale_edge_alpha(range = c(0.4, 0.9)) +
  ggraph::scale_edge_colour_manual(
    values = c(
      positive = "steelblue4",
      negative = "firebrick3"
    )
  ) +
  ggraph::scale_edge_linetype_manual(
    values = c(
      positive = "solid",
      negative = "dashed"
    )
  ) +
  ggplot2::labs(
    title = "Root subset mini-network: pathogen, root disease, foliar disease, and yield"
  ) +
  ggplot2::theme_void()

ggplot2::ggsave(
  "fig_root_mini_network.png",
  p_root_net,
  width = 9,
  height = 6,
  dpi = 300
)
```

# -----------------------------
# 4. PCA for root subset
# -----------------------------
```{r}
# Numeric PCA input
root_pca_df <- df_root %>%
  select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  ) %>%
  drop_na()

root_pca_res <- FactoMineR::PCA(root_pca_df, scale.unit = TRUE, graph = FALSE)

# Save loadings
root_pca_loadings <- as.data.frame(root_pca_res$var$coord) %>%
  rownames_to_column("variable")
write_csv(root_pca_loadings, "root_pca_loadings.csv")

# Scree plot
p_root_scree <- factoextra::fviz_eig(root_pca_res, addlabels = TRUE) +
  labs(title = "Root subset PCA scree plot")

ggsave("fig_root_pca_scree.png", p_root_scree, width = 7, height = 5, dpi = 300)

# PCA biplot by condensed Fv bin
root_pca_fv <- df_root %>%
  select(
    fv_level_simple,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  ) %>%
  drop_na()

root_pca_res_fv <- FactoMineR::PCA(
  root_pca_fv %>% select(-fv_level_simple),
  scale.unit = TRUE,
  graph = FALSE
)

p_root_pca_fv <- factoextra::fviz_pca_biplot(
  root_pca_res_fv,
  geom.ind = "point",
  habillage = root_pca_fv$fv_level_simple,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  labs(title = "Root subset PCA biplot by Fv abundance class")

ggsave("fig_root_pca_biplot_fv_simple.png", p_root_pca_fv, width = 8, height = 6, dpi = 300)

# Optional finer Fv factor version
root_pca_fv_simple <- df_root %>%
  select(
    fv_level_simple,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  ) %>%
  drop_na()

root_pca_res_fv_simple <- FactoMineR::PCA(
  root_pca_fv_simple %>% select(-fv_level_simple),
  scale.unit = TRUE,
  graph = FALSE
)

p_root_pca_fv_simple <- factoextra::fviz_pca_biplot(
  root_pca_res_fv_simple,
  geom.ind = "point",
  habillage = root_pca_fv_simple$fv_level_simple,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  labs(title = "Root subset PCA biplot by detailed Fv abundance class")

ggsave("fig_root_pca_biplot_fv_simple.png", p_root_pca_fv_simple, width = 9, height = 7, dpi = 300)

# PCA by product
root_pca_prod <- df_root %>%
  select(
    product,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  ) %>%
  drop_na()

root_pca_res_prod <- FactoMineR::PCA(
  root_pca_prod %>% select(-product),
  scale.unit = TRUE,
  graph = FALSE
)

p_root_pca_prod <- factoextra::fviz_pca_biplot(
  root_pca_res_prod,
  geom.ind = "point",
  habillage = root_pca_prod$product,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  labs(title = "Root subset PCA biplot by product")

ggsave("fig_root_pca_biplot_product.png", p_root_pca_prod, width = 8, height = 6, dpi = 300)

# PCA by SDS tolerance
root_pca_tol <- df_root %>%
  select(
    sds_tol,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    rdx,
    drw,
    fdx,
    yield_kgha
  ) %>%
  drop_na()

root_pca_res_tol <- FactoMineR::PCA(
  root_pca_tol %>% select(-sds_tol),
  scale.unit = TRUE,
  graph = FALSE
)

p_root_pca_tol <- factoextra::fviz_pca_biplot(
  root_pca_res_tol,
  geom.ind = "point",
  habillage = root_pca_tol$sds_tol,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  labs(title = "Root subset PCA biplot by SDS tolerance")

ggsave("fig_root_pca_biplot_sds_tol.png", p_root_pca_tol, width = 8, height = 6, dpi = 300)
```
# -----------------------------
# 6A. FDX model
# -----------------------------
```{r}
# Type III ANOVA requires appropriate contrasts
options(contrasts = c("contr.sum", "contr.poly"))

# Build model-ready datasets with consistent variables
df_fdx_mod <- df_fdx %>%
  mutate(
    # scale continuous predictors for interaction stability / interpretability
    z_log10_fv  = as.numeric(scale(log10_fv)),
    z_log10_scn = as.numeric(scale(log10_scn)),
    exp_siteyear = interaction(siteyear, exp, drop = TRUE)
  )

df_fdx_ph_mod <- df_fdx_ph %>%
  mutate(
    z_log10_fv  = as.numeric(scale(log10_fv)),
    z_log10_scn = as.numeric(scale(log10_scn)),
    z_soil_ph   = as.numeric(scale(soil_ph)),
    exp_siteyear = interaction(siteyear, exp, drop = TRUE)
  )

# Core model
# state dropped because siteyear already captures environment
# exp nested within siteyear via exp_siteyear
m_fdx_1 <- lmer(
  log1p_fdx ~ z_log10_fv + z_log10_scn + product + sds_tol +
  (1 | siteyear) + (1 | exp_siteyear),
  data = df_fdx_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

cat("\n================ FDX model 1 ================\n")
print(summary(m_fdx_1))
print(joint_tests(m_fdx_1))
print(performance::check_singularity(m_fdx_1))
print(performance::check_collinearity(m_fdx_1))

# Optional pH model
m_fdx_2 <- lmer(
  log1p_fdx ~ z_log10_fv + z_log10_scn + log1p_rdx + z_soil_ph + product:sds_tol +
  (1 | siteyear) + (1 | exp_siteyear),
  data = df_fdx_ph_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

cat("\n================ FDX model 2 (with pH) ================\n")
print(summary(m_fdx_2))
print(joint_tests(m_fdx_2))
print(performance::check_singularity(m_fdx_2))
print(performance::check_collinearity(m_fdx_2))

# IMPORTANT:
# To compare AIC, fit the no-pH model on the same rows used by the pH model
m_fdx_1_same_rows <- lmer(
  log1p_fdx ~ z_log10_fv + z_log10_scn + product + sds_tol +
  (1 | siteyear) + (1 | exp_siteyear),
  data = df_fdx_ph_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

cat("\n================ AIC comparison on same rows ================\n")
print(AIC(m_fdx_1_same_rows, m_fdx_2))

# EMMs for product from core model
emm_fdx_product <- emmeans(m_fdx_2, ~ product:sds_tol, type = "response")

multcomp::cld(emm_fdx_product)

cat("\n================ FDX product EMMs ================\n")
summary(m_fdx_2)
print(multcomp::cld(emm_fdx_product))
print(pairs(emm_fdx_product, adjust = "tukey"))
print(cld(emm_fdx_product, Letters = letters, adjust = "tukey"))


fdx_emm_bt <- as.data.frame(emm_fdx_product) %>%
  dplyr::mutate(
    emmean_bt   = exp(emmean) - 1,
    lower.CL_bt = exp(lower.CL) - 1,
    upper.CL_bt = exp(upper.CL) - 1
  )

fdx_emm_bt_fmt <- fdx_emm_bt %>%
  dplyr::transmute(
    product:sds_tol,
    EMM_bt = round(emmean_bt, 2),
    lower.CL_bt = round(lower.CL_bt, 2),
    upper.CL_bt = round(upper.CL_bt, 2)
  )

#emmeans::regrid(emm_fdx_product, transform = "response")
#performance::check_singularity(m_fdx_1)
#performance::check_singularity(m_fdx_2)
```

# ============================================================
# FDX downstream analyses
# Correlation, association matrix, mini-network, PCA
# ============================================================

# -----------------------------
# 1. FDX subset numeric correlations
# -----------------------------
```{r}
fdx_num <- df_fdx %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  )

# Spearman correlation matrix
fdx_cor <- stats::cor(
  fdx_num,
  method = "spearman",
  use = "pairwise.complete.obs"
)

cat("\n================ FDX subset numeric Spearman correlation matrix ================\n")
print(fdx_cor)

# Pairwise Spearman tests
fdx_cor_tests <- expand.grid(
  var1 = names(fdx_num),
  var2 = names(fdx_num),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(var1 < var2) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    test = list(
      suppressWarnings(
        stats::cor.test(
          fdx_num[[var1]],
          fdx_num[[var2]],
          method = "spearman",
          use = "pairwise.complete.obs"
        )
      )
    ),
    rho = unname(test$estimate),
    p_value = test$p.value
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(var1, var2, rho, p_value)

cat("\n================ FDX subset pairwise Spearman correlations ================\n")
print(fdx_cor_tests)

readr::write_csv(fdx_cor_tests, "fdx_pairwise_spearman_correlations.csv")

# Visualize numeric correlation matrix
png("fig_fdx_numeric_correlation_matrix.png", width = 1500, height = 1300, res = 220)
corrplot::corrplot(
  fdx_cor,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.75,
  mar = c(0, 0, 2, 0),
  title = "FDX subset numeric Spearman correlation matrix"
)
dev.off()

```

# -----------------------------
# 2. Mixed-type association matrix for FDX subset
# -----------------------------
```{r}
# Helper functions (safe versions)
safe_extract_result <- function(res) {
  out <- c(estimate = NA_real_, p = NA_real_)

  if (is.null(res) || length(res) == 0) return(out)

  if (is.numeric(res) && length(res) == 1 && is.null(names(res))) {
    out["estimate"] <- as.numeric(res)
    return(out)
  }

  if (is.atomic(res) && !is.null(names(res))) {
    if ("estimate" %in% names(res) && length(res[["estimate"]]) == 1) {
      out["estimate"] <- suppressWarnings(as.numeric(res[["estimate"]]))
    }
    if ("p" %in% names(res) && length(res[["p"]]) == 1) {
      out["p"] <- suppressWarnings(as.numeric(res[["p"]]))
    }
    return(out)
  }

  if (is.list(res)) {
    if (!is.null(res$estimate) && length(res$estimate) == 1) {
      out["estimate"] <- suppressWarnings(as.numeric(res$estimate))
    }
    if (!is.null(res$p) && length(res$p) == 1) {
      out["p"] <- suppressWarnings(as.numeric(res$p))
    }
    return(out)
  }

  out
}

safe_cor_test <- function(x, y, method = "pearson") {
  ok <- stats::complete.cases(x, y)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  out <- tryCatch(
    suppressWarnings(stats::cor.test(x[ok], y[ok], method = method)),
    error = function(e) NULL
  )

  if (is.null(out)) return(c(estimate = NA_real_, p = NA_real_))

  c(
    estimate = unname(out$estimate),
    p = out$p.value
  )
}

safe_eta <- function(num, fac) {
  ok <- stats::complete.cases(num, fac)
  if (sum(ok) < 5 || nlevels(factor(fac[ok])) < 2) {
    return(c(estimate = NA_real_, p = NA_real_))
  }

  fit <- tryCatch(
    stats::aov(num[ok] ~ factor(fac[ok])),
    error = function(e) NULL
  )

  if (is.null(fit)) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    DescTools::EtaSq(fit, type = 1)[1, "eta.sq"],
    error = function(e) NA_real_
  )

  p <- tryCatch(
    summary(fit)[[1]][["Pr(>F)"]][1],
    error = function(e) NA_real_
  )

  c(estimate = est, p = p)
}

safe_cramers_v <- function(x, y) {
  ok <- stats::complete.cases(x, y)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  tab <- table(x[ok], y[ok])
  if (min(dim(tab)) < 2) return(c(estimate = NA_real_, p = NA_real_))

  chi <- tryCatch(
    suppressWarnings(stats::chisq.test(tab)),
    error = function(e) NULL
  )

  est <- tryCatch(
    DescTools::CramerV(tab),
    error = function(e) NA_real_
  )

  p <- if (is.null(chi)) NA_real_ else chi$p.value

  c(estimate = est, p = p)
}

safe_polyserial <- function(x, y_ord) {
  ok <- stats::complete.cases(x, y_ord)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    polycor::polyserial(x[ok], as.ordered(y_ord[ok]), std.err = FALSE),
    error = function(e) NA_real_
  )

  c(estimate = est, p = NA_real_)
}

safe_polychor <- function(x_ord, y_ord) {
  ok <- stats::complete.cases(x_ord, y_ord)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    polycor::polychor(x_ord[ok], y_ord[ok]),
    error = function(e) NA_real_
  )

  c(estimate = est, p = NA_real_)
}

fdx_assoc_df <- df_fdx %>%
  dplyr::transmute(
    ratio_fvfg_dnang = ratio_fvfg_dnang,
    scn_count_rep    = scn_count_rep,
    soil_ph          = soil_ph,
    fdx              = fdx,
    yield_kgha       = yield_kgha,
    product          = as.factor(product),
    sds_tol          = as.factor(sds_tol),
    fdx_level        = factor(fdx_level, ordered = TRUE),
    scn_level        = factor(scn_level, ordered = TRUE),
    fv_level_simple  = factor(fv_level_simple, ordered = TRUE)
  )

fdx_var_names <- names(fdx_assoc_df)

fdx_var_type <- c(
  ratio_fvfg_dnang = "numeric",
  scn_count_rep    = "numeric",
  soil_ph          = "numeric",
  fdx              = "numeric",
  yield_kgha       = "numeric",
  product          = "nominal",
  sds_tol          = "nominal",
  fdx_level        = "ordinal",
  scn_level        = "ordinal",
  fv_level_simple  = "ordinal"
)

fdx_assoc_mat <- matrix(
  NA_real_,
  nrow = length(fdx_var_names),
  ncol = length(fdx_var_names),
  dimnames = list(fdx_var_names, fdx_var_names)
)

fdx_p_mat <- matrix(
  NA_real_,
  nrow = length(fdx_var_names),
  ncol = length(fdx_var_names),
  dimnames = list(fdx_var_names, fdx_var_names)
)

for (i in seq_along(fdx_var_names)) {
  for (j in seq_along(fdx_var_names)) {

    xi <- fdx_assoc_df[[fdx_var_names[i]]]
    xj <- fdx_assoc_df[[fdx_var_names[j]]]
    ti <- fdx_var_type[[fdx_var_names[i]]]
    tj <- fdx_var_type[[fdx_var_names[j]]]

    if (i == j) {
      fdx_assoc_mat[i, j] <- 1
      fdx_p_mat[i, j] <- 0
      next
    }

    res <- c(estimate = NA_real_, p = NA_real_)

    if (ti == "numeric" && tj == "numeric") {
      res <- safe_cor_test(xi, xj, method = "pearson")
    } else if (ti == "numeric" && tj == "nominal") {
      res <- safe_eta(xi, xj)
    } else if (ti == "nominal" && tj == "numeric") {
      res <- safe_eta(xj, xi)
    } else if (ti == "nominal" && tj == "nominal") {
      res <- safe_cramers_v(xi, xj)
    } else if (ti == "numeric" && tj == "ordinal") {
      res <- safe_polyserial(xi, xj)
    } else if (ti == "ordinal" && tj == "numeric") {
      res <- safe_polyserial(xj, xi)
    } else if (ti == "ordinal" && tj == "ordinal") {
      res <- safe_polychor(xi, xj)
    } else if (ti == "nominal" && tj == "ordinal") {
      res <- safe_cramers_v(xi, xj)
    } else if (ti == "ordinal" && tj == "nominal") {
      res <- safe_cramers_v(xi, xj)
    }

    res_clean <- safe_extract_result(res)

    est_val <- if (length(res_clean["estimate"]) == 1) as.numeric(res_clean["estimate"]) else NA_real_
    p_val   <- if (length(res_clean["p"]) == 1) as.numeric(res_clean["p"]) else NA_real_

    fdx_assoc_mat[i, j] <- est_val
    fdx_p_mat[i, j] <- p_val
  }
}

readr::write_csv(
  as.data.frame(fdx_assoc_mat) %>% tibble::rownames_to_column("variable"),
  "fdx_association_matrix.csv"
)

readr::write_csv(
  as.data.frame(fdx_p_mat) %>% tibble::rownames_to_column("variable"),
  "fdx_association_pvalues.csv"
)

png("fig_fdx_association_matrix.png", width = 1600, height = 1450, res = 220)
corrplot::corrplot(
  fdx_assoc_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  na.label = " ",
  mar = c(0, 0, 2, 0),
  title = "FDX subset mixed-type association matrix"
)
dev.off()

```
# -----------------------------
# 3. Mini-network association mapping
# -----------------------------
```{r}
fdx_net_df <- df_fdx %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_net_cor <- stats::cor(
  fdx_net_df,
  method = "spearman",
  use = "pairwise.complete.obs"
)

fdx_net_p <- matrix(
  NA_real_,
  nrow = ncol(fdx_net_df),
  ncol = ncol(fdx_net_df),
  dimnames = list(names(fdx_net_df), names(fdx_net_df))
)

for (i in seq_len(ncol(fdx_net_df))) {
  for (j in seq_len(ncol(fdx_net_df))) {
    if (i == j) {
      fdx_net_p[i, j] <- 0
    } else {
      ct <- suppressWarnings(
        stats::cor.test(
          fdx_net_df[[i]],
          fdx_net_df[[j]],
          method = "spearman"
        )
      )
      fdx_net_p[i, j] <- ct$p.value
    }
  }
}

fdx_edge_tbl <- as.data.frame(as.table(fdx_net_cor)) %>%
  dplyr::rename(from = Var1, to = Var2, weight = Freq) %>%
  dplyr::mutate(p = as.vector(fdx_net_p)) %>%
  dplyr::filter(from != to) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(pair = paste(sort(c(from, to)), collapse = "__")) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(pair, .keep_all = TRUE) %>%
  dplyr::filter(!is.na(weight), abs(weight) >= 0.10, p < 0.05) %>%
  dplyr::mutate(
    layout_weight = abs(weight),
    sign_class = factor(ifelse(weight > 0, "positive", "negative"),
                        levels = c("positive", "negative"))
  )

fdx_node_tbl <- tibble::tibble(
  name = colnames(fdx_net_cor),
  category = factor(
    c("pathogen", "pathogen", "soil", "disease", "yield"),
    levels = c("pathogen", "soil", "disease", "yield")
  )
)

fdx_graph_obj <- tidygraph::tbl_graph(
  nodes = fdx_node_tbl,
  edges = fdx_edge_tbl,
  directed = FALSE
)

p_fdx_net <- ggraph::ggraph(
  fdx_graph_obj,
  layout = "fr",
  weights = layout_weight
) +
  ggraph::geom_edge_link(
    ggplot2::aes(
      edge_width = abs(weight),
      edge_alpha = abs(weight),
      edge_colour = sign_class,
      edge_linetype = sign_class
    )
  ) +
  ggraph::geom_node_point(
    ggplot2::aes(shape = category),
    size = 8
  ) +
  ggraph::geom_node_text(
    ggplot2::aes(label = name),
    repel = TRUE
  ) +
  ggraph::scale_edge_width(range = c(0.5, 2.5)) +
  ggraph::scale_edge_alpha(range = c(0.4, 0.9)) +
  ggraph::scale_edge_colour_manual(
    values = c(
      positive = "steelblue4",
      negative = "firebrick3"
    )
  ) +
  ggraph::scale_edge_linetype_manual(
    values = c(
      positive = "solid",
      negative = "dashed"
    )
  ) +
  ggplot2::labs(
    title = "FDX subset mini-network: pathogen, foliar disease, soil, and yield"
  ) +
  ggplot2::theme_void()

ggplot2::ggsave(
  "fig_fdx_mini_network.png",
  p_fdx_net,
  width = 8,
  height = 6,
  dpi = 300
)

```

# -----------------------------
# 4. PCA for FDX subset
# -----------------------------
```{r}
fdx_pca_df <- df_fdx %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_pca_res <- FactoMineR::PCA(fdx_pca_df, scale.unit = TRUE, graph = FALSE)

# Save loadings
fdx_pca_loadings <- as.data.frame(fdx_pca_res$var$coord) %>%
  tibble::rownames_to_column("variable")
readr::write_csv(fdx_pca_loadings, "fdx_pca_loadings.csv")

# Scree plot
p_fdx_scree <- factoextra::fviz_eig(fdx_pca_res, addlabels = TRUE) +
  ggplot2::labs(title = "FDX subset PCA scree plot")

ggplot2::ggsave("fig_fdx_pca_scree.png", p_fdx_scree, width = 7, height = 5, dpi = 300)

# PCA biplot by FDX level
fdx_pca_fdxlvl <- df_fdx %>%
  dplyr::select(
    fdx_level,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_pca_res_fdxlvl <- FactoMineR::PCA(
  fdx_pca_fdxlvl %>% dplyr::select(-fdx_level),
  scale.unit = TRUE,
  graph = FALSE
)

p_fdx_pca_fdxlvl <- factoextra::fviz_pca_biplot(
  fdx_pca_res_fdxlvl,
  geom.ind = "point",
  habillage = fdx_pca_fdxlvl$fdx_level,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  ggplot2::labs(title = "FDX subset PCA biplot by FDX level")

ggplot2::ggsave("fig_fdx_pca_biplot_fdx_level.png", p_fdx_pca_fdxlvl, width = 8, height = 6, dpi = 300)

# PCA biplot by SCN level
fdx_pca_scn <- df_fdx %>%
  dplyr::select(
    scn_level,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_pca_res_scn <- FactoMineR::PCA(
  fdx_pca_scn %>% dplyr::select(-scn_level),
  scale.unit = TRUE,
  graph = FALSE
)

p_fdx_pca_scn <- factoextra::fviz_pca_biplot(
  fdx_pca_res_scn,
  geom.ind = "point",
  habillage = fdx_pca_scn$scn_level,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  ggplot2::labs(title = "FDX subset PCA biplot by SCN level")

ggplot2::ggsave("fig_fdx_pca_biplot_scn_level.png", p_fdx_pca_scn, width = 8, height = 6, dpi = 300)

# PCA biplot by condensed Fv level
fdx_pca_fv_simple <- df_fdx %>%
  dplyr::select(
    fv_level_simple,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_pca_res_fv_simple <- FactoMineR::PCA(
  fdx_pca_fv_simple %>% dplyr::select(-fv_level_simple),
  scale.unit = TRUE,
  graph = FALSE
)

p_fdx_pca_fv_simple <- factoextra::fviz_pca_biplot(
  fdx_pca_res_fv_simple,
  geom.ind = "point",
  habillage = fdx_pca_fv_simple$fv_level_simple,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  ggplot2::labs(title = "FDX subset PCA biplot by Fv abundance class")

ggplot2::ggsave("fig_fdx_pca_biplot_fv_simple.png", p_fdx_pca_fv_simple, width = 8, height = 6, dpi = 300)

# PCA biplot by detailed Fv level
fdx_pca_fv_simple <- df_fdx %>%
  dplyr::select(
    fv_level_simple,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_pca_res_fv_simple <- FactoMineR::PCA(
  fdx_pca_fv_simple %>% dplyr::select(-fv_level_simple),
  scale.unit = TRUE,
  graph = FALSE
)

p_fdx_pca_fv_simple <- factoextra::fviz_pca_biplot(
  fdx_pca_res_fv_simple,
  geom.ind = "point",
  habillage = fdx_pca_fv_simple$fv_level_simple,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  ggplot2::labs(title = "FDX subset PCA biplot by detailed Fv abundance class")

ggplot2::ggsave("fig_fdx_pca_biplot_fv_simple.png", p_fdx_pca_fv_simple, width = 9, height = 7, dpi = 300)

# PCA biplot by product
fdx_pca_prod <- df_fdx %>%
  dplyr::select(
    product,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

fdx_pca_res_prod <- FactoMineR::PCA(
  fdx_pca_prod %>% dplyr::select(-product),
  scale.unit = TRUE,
  graph = FALSE
)

p_fdx_pca_prod <- factoextra::fviz_pca_biplot(
  fdx_pca_res_prod,
  geom.ind = "point",
  habillage = fdx_pca_prod$product,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  repel = TRUE
) +
  ggplot2::labs(title = "FDX subset PCA biplot by product")

ggplot2::ggsave("fig_fdx_pca_biplot_product.png", p_fdx_pca_prod, width = 8, height = 6, dpi = 300)
```
# -----------------------------
# 6B. Yield LMMs
# -----------------------------
```{r}

tmp <- df_yield %>%
  summarise(
    n_siteyear = n_distinct(siteyear),
    n_state    = n_distinct(state),
    n_location = n_distinct(location),
    n_year     = n_distinct(year)
  )

cat(
  paste0(
    "Field trials were conducted across ",
    tmp$n_siteyear, " site-years spanning ",
    tmp$n_state, " U.S. states and Ontario, Canada, encompassing ",
    tmp$n_location, " locations over ",
    tmp$n_year, " growing seasons."
  )
)

# Build model-ready datasets with consistent variables
df_yield_mod <- df_yield %>%
  mutate(
    # scale continuous predictors for interaction stability / interpretability
    z_log10_fv  = as.numeric(scale(log10_fv)),
    z_log10_scn = as.numeric(scale(log10_scn)),
    z_soil_ph   = as.numeric(scale(soil_ph)),
    exp_siteyear = interaction(siteyear, exp, drop = TRUE)
  )

cat("\n================ Yield model 1: pathogen + environment ================\n")

m_yield_1 <- lmer(
  yield_kgha ~
    z_log10_fv + 
    z_log10_scn +
    log1p_fdx +
    log1p_rdx +
    z_soil_ph +
    sds_tol:product +
    (1 | siteyear) +
    (1 | exp_siteyear),
  data = df_yield_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

print(summary(m_yield_1))
print(emmeans::joint_tests(m_yield_1))
print(multcomp::cld(emmeans(m_yield_1, ~ sds_tol)))

cat("\nChecking singularity:\n")
print(performance::check_singularity(m_yield_1))

cat("\nChecking multicollinearity:\n")
print(performance::check_collinearity(m_yield_1))


# ================================
# Yield model 2: add foliar SDS
# ================================

cat("\n================ Yield model 2: + FDX ================\n")

m_yield_2 <- lmer(
  yield_kgha ~
    z_log10_fv +
    z_log10_scn +
    z_soil_ph +
    product:sds_tol +
    (1|log1p_fdx) +
    (1|log1p_rdx) +
    (1 | siteyear) +
    (1 | exp_siteyear),
  data = df_yield_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

print(summary(m_yield_2))
print(emmeans::joint_tests(m_yield_2))

cat("\nChecking singularity:\n")
print(performance::check_singularity(m_yield_2))

cat("\nChecking multicollinearity:\n")
print(performance::check_collinearity(m_yield_2))


# ================================
# Yield model 3: add root disease
# ================================

cat("\n================ Yield model 3: + RDX ================\n")

m_yield_3 <- lmer(
  yield_kgha ~
    log1p_fdx +
    log1p_rdx +
    product +
    (1 | siteyear) +
    (1 | exp_siteyear),
  data = df_yield_mod,
  REML = FALSE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 200000))
)

print(summary(m_yield_3))
print(emmeans::joint_tests(m_yield_3))

cat("\nChecking singularity:\n")
print(performance::check_singularity(m_yield_3))

cat("\nChecking multicollinearity:\n")
print(performance::check_collinearity(m_yield_3))


# ================================
# Model comparison
# ================================

cat("\n================ Yield model comparison ================\n")

print(AIC(m_yield_1, m_yield_2, m_yield_3))


# ================================
# Product marginal means
# ================================

emm_yield_product <- emmeans(m_yield_2, ~ product:sds_tol)

cat("\n================ Yield product EMMs ================\n")

print(emm_yield_product)
print(pairs(emm_yield_product, adjust = "tukey"))

print(
  multcomp::cld(
    emm_yield_product,
    Letters = letters,
    adjust = "tukey"
  )
)
```
# ============================================================
# Yield downstream analyses
# Correlation, association matrix, mini-network, PCA
# ============================================================

# -----------------------------
# 1. Yield subset numeric correlations
# -----------------------------
```{r}
yield_num <- df_yield_mod %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

yield_cor <- stats::cor(
  yield_num,
  method = "spearman",
  use = "pairwise.complete.obs"
)

cat("\n================ Yield subset numeric Spearman correlation matrix ================\n")
print(yield_cor)

yield_cor_tests <- expand.grid(
  var1 = names(yield_num),
  var2 = names(yield_num),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(var1 < var2) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    test = list(
      suppressWarnings(
        stats::cor.test(
          yield_num[[var1]],
          yield_num[[var2]],
          method = "spearman",
          use = "pairwise.complete.obs"
        )
      )
    ),
    rho = unname(test$estimate),
    p_value = test$p.value
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(var1, var2, rho, p_value)

cat("\n================ Yield subset pairwise Spearman correlations ================\n")
print(yield_cor_tests)

readr::write_csv(yield_cor_tests, "yield_pairwise_spearman_correlations.csv")

png("fig_yield_numeric_correlation_matrix.png", width = 1600, height = 1400, res = 220)
corrplot::corrplot(
  yield_cor,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  addCoef.col = "black",
  number.cex = 0.7,
  mar = c(0, 0, 2, 0),
  title = "Yield subset numeric Spearman correlation matrix"
)
dev.off()

```

# -----------------------------
# 2. Mixed-type association matrix
# -----------------------------
```{r}
# safe helpers reused here for self-contained chunk

safe_extract_result <- function(res) {
  out <- c(estimate = NA_real_, p = NA_real_)

  if (is.null(res) || length(res) == 0) return(out)

  if (is.numeric(res) && length(res) == 1 && is.null(names(res))) {
    out["estimate"] <- as.numeric(res)
    return(out)
  }

  if (is.atomic(res) && !is.null(names(res))) {
    if ("estimate" %in% names(res) && length(res[["estimate"]]) == 1) {
      out["estimate"] <- suppressWarnings(as.numeric(res[["estimate"]]))
    }
    if ("p" %in% names(res) && length(res[["p"]]) == 1) {
      out["p"] <- suppressWarnings(as.numeric(res[["p"]]))
    }
    return(out)
  }

  if (is.list(res)) {
    if (!is.null(res$estimate) && length(res$estimate) == 1) {
      out["estimate"] <- suppressWarnings(as.numeric(res$estimate))
    }
    if (!is.null(res$p) && length(res$p) == 1) {
      out["p"] <- suppressWarnings(as.numeric(res$p))
    }
    return(out)
  }

  out
}

safe_cor_test <- function(x, y, method = "pearson") {
  ok <- stats::complete.cases(x, y)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  out <- tryCatch(
    suppressWarnings(stats::cor.test(x[ok], y[ok], method = method)),
    error = function(e) NULL
  )

  if (is.null(out)) return(c(estimate = NA_real_, p = NA_real_))

  c(
    estimate = unname(out$estimate),
    p = out$p.value
  )
}

safe_eta <- function(num, fac) {
  ok <- stats::complete.cases(num, fac)
  if (sum(ok) < 5 || nlevels(factor(fac[ok])) < 2) {
    return(c(estimate = NA_real_, p = NA_real_))
  }

  fit <- tryCatch(
    stats::aov(num[ok] ~ factor(fac[ok])),
    error = function(e) NULL
  )

  if (is.null(fit)) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    DescTools::EtaSq(fit, type = 1)[1, "eta.sq"],
    error = function(e) NA_real_
  )

  p <- tryCatch(
    summary(fit)[[1]][["Pr(>F)"]][1],
    error = function(e) NA_real_
  )

  c(estimate = est, p = p)
}

safe_cramers_v <- function(x, y) {
  ok <- stats::complete.cases(x, y)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  tab <- table(x[ok], y[ok])
  if (min(dim(tab)) < 2) return(c(estimate = NA_real_, p = NA_real_))

  chi <- tryCatch(
    suppressWarnings(stats::chisq.test(tab)),
    error = function(e) NULL
  )

  est <- tryCatch(
    DescTools::CramerV(tab),
    error = function(e) NA_real_
  )

  p <- if (is.null(chi)) NA_real_ else chi$p.value

  c(estimate = est, p = p)
}

safe_polyserial <- function(x, y_ord) {
  ok <- stats::complete.cases(x, y_ord)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    polycor::polyserial(x[ok], as.ordered(y_ord[ok]), std.err = FALSE),
    error = function(e) NA_real_
  )

  c(estimate = est, p = NA_real_)
}

safe_polychor <- function(x_ord, y_ord) {
  ok <- stats::complete.cases(x_ord, y_ord)
  if (sum(ok) < 5) return(c(estimate = NA_real_, p = NA_real_))

  est <- tryCatch(
    polycor::polychor(x_ord[ok], y_ord[ok]),
    error = function(e) NA_real_
  )

  c(estimate = est, p = NA_real_)
}

yield_assoc_df <- df_fdx_ph_mod %>%
  dplyr::transmute(
    ratio_fvfg_dnang = ratio_fvfg_dnang,
    scn_count_rep    = scn_count_rep,
    soil_ph          = soil_ph,
    fdx              = fdx,
    rdx              = rdx,
    yield_kgha       = yield_kgha,
    product          = as.factor(product),
    sds_tol          = as.factor(sds_tol),
    fdx_level        = factor(fdx_level, ordered = TRUE),
    scn_level        = factor(scn_level, ordered = TRUE),
    fv_level_simple  = factor(fv_level_simple, ordered = TRUE)
  )

yield_var_names <- names(yield_assoc_df)

yield_var_type <- c(
  ratio_fvfg_dnang = "numeric",
  scn_count_rep    = "numeric",
  soil_ph          = "numeric",
  fdx              = "numeric",
  rdx              = "numeric",
  yield_kgha       = "numeric",
  product          = "nominal",
  sds_tol          = "nominal",
  fdx_level        = "ordinal",
  scn_level        = "ordinal",
  fv_level_simple  = "ordinal"
)

yield_assoc_mat <- matrix(
  NA_real_,
  nrow = length(yield_var_names),
  ncol = length(yield_var_names),
  dimnames = list(yield_var_names, yield_var_names)
)

yield_p_mat <- matrix(
  NA_real_,
  nrow = length(yield_var_names),
  ncol = length(yield_var_names),
  dimnames = list(yield_var_names, yield_var_names)
)

for (i in seq_along(yield_var_names)) {
  for (j in seq_along(yield_var_names)) {

    xi <- yield_assoc_df[[yield_var_names[i]]]
    xj <- yield_assoc_df[[yield_var_names[j]]]
    ti <- yield_var_type[[yield_var_names[i]]]
    tj <- yield_var_type[[yield_var_names[j]]]

    if (i == j) {
      yield_assoc_mat[i, j] <- 1
      yield_p_mat[i, j] <- 0
      next
    }

    res <- c(estimate = NA_real_, p = NA_real_)

    if (ti == "numeric" && tj == "numeric") {
      res <- safe_cor_test(xi, xj, method = "pearson")
    } else if (ti == "numeric" && tj == "nominal") {
      res <- safe_eta(xi, xj)
    } else if (ti == "nominal" && tj == "numeric") {
      res <- safe_eta(xj, xi)
    } else if (ti == "nominal" && tj == "nominal") {
      res <- safe_cramers_v(xi, xj)
    } else if (ti == "numeric" && tj == "ordinal") {
      res <- safe_polyserial(xi, xj)
    } else if (ti == "ordinal" && tj == "numeric") {
      res <- safe_polyserial(xj, xi)
    } else if (ti == "ordinal" && tj == "ordinal") {
      res <- safe_polychor(xi, xj)
    } else if (ti == "nominal" && tj == "ordinal") {
      res <- safe_cramers_v(xi, xj)
    } else if (ti == "ordinal" && tj == "nominal") {
      res <- safe_cramers_v(xi, xj)
    }

    res_clean <- safe_extract_result(res)

    est_val <- if (length(res_clean["estimate"]) == 1) as.numeric(res_clean["estimate"]) else NA_real_
    p_val   <- if (length(res_clean["p"]) == 1) as.numeric(res_clean["p"]) else NA_real_

    yield_assoc_mat[i, j] <- est_val
    yield_p_mat[i, j] <- p_val
  }
}

readr::write_csv(
  as.data.frame(yield_assoc_mat) %>% tibble::rownames_to_column("variable"),
  "yield_association_matrix.csv"
)

readr::write_csv(
  as.data.frame(yield_p_mat) %>% tibble::rownames_to_column("variable"),
  "yield_association_pvalues.csv"
)

png("fig_yield_association_matrix.png", width = 1700, height = 1500, res = 220)
corrplot::corrplot(
  yield_assoc_mat,
  method = "color",
  type = "upper",
  tl.col = "black",
  tl.srt = 45,
  na.label = " ",
  mar = c(0, 0, 2, 0),
  title = "Yield subset mixed-type association matrix"
)
dev.off()

```
# -----------------------------
# 3. Mini-network association mapping
# -----------------------------
```{r}
yield_net_df <- df_fdx_ph_mod %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na()

yield_net_cor <- stats::cor(
  yield_net_df,
  method = "spearman",
  use = "pairwise.complete.obs"
)

yield_net_p <- matrix(
  NA_real_,
  nrow = ncol(yield_net_df),
  ncol = ncol(yield_net_df),
  dimnames = list(names(yield_net_df), names(yield_net_df))
)

for (i in seq_len(ncol(yield_net_df))) {
  for (j in seq_len(ncol(yield_net_df))) {
    if (i == j) {
      yield_net_p[i, j] <- 0
    } else {
      ct <- suppressWarnings(
        stats::cor.test(
          yield_net_df[[i]],
          yield_net_df[[j]],
          method = "spearman"
        )
      )
      yield_net_p[i, j] <- ct$p.value
    }
  }
}

yield_edge_tbl <- as.data.frame(as.table(yield_net_cor)) %>%
  dplyr::rename(from = Var1, to = Var2, weight = Freq) %>%
  dplyr::mutate(p = as.vector(yield_net_p)) %>%
  dplyr::filter(from != to) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(pair = paste(sort(c(from, to)), collapse = "__")) %>%
  dplyr::ungroup() %>%
  dplyr::distinct(pair, .keep_all = TRUE) %>%
  dplyr::filter(!is.na(weight), abs(weight) >= 0.10, p < 0.05) %>%
  dplyr::mutate(
    layout_weight = abs(weight),
    sign_class = factor(ifelse(weight > 0, "positive", "negative"),
                        levels = c("positive", "negative"))
  )

yield_node_tbl <- tibble::tibble(
  name = colnames(yield_net_cor),
  category = factor(
    c("pathogen", "pathogen", "soil", "disease", "root", "yield"),
    levels = c("pathogen", "soil", "root", "disease", "yield")
  )
)

yield_graph_obj <- tidygraph::tbl_graph(
  nodes = yield_node_tbl,
  edges = yield_edge_tbl,
  directed = FALSE
)

p_yield_net <- ggraph::ggraph(
  yield_graph_obj,
  layout = "fr",
  weights = layout_weight
) +
  ggraph::geom_edge_link(
    ggplot2::aes(
      edge_width = abs(weight),
      edge_alpha = abs(weight),
      edge_colour = sign_class,
      edge_linetype = sign_class
    )
  ) +
  ggraph::geom_node_point(
    ggplot2::aes(shape = category),
    size = 8
  ) +
  ggraph::geom_node_text(
    ggplot2::aes(label = name),
    repel = TRUE
  ) +
  ggraph::scale_edge_width(range = c(0.5, 2.5)) +
  ggraph::scale_edge_alpha(range = c(0.4, 0.9)) +
  ggraph::scale_edge_colour_manual(
    values = c(
      positive = "steelblue4",
      negative = "firebrick3"
    )
  ) +
  ggraph::scale_edge_linetype_manual(
    values = c(
      positive = "solid",
      negative = "dashed"
    )
  ) +
  ggplot2::labs(
    title = "Yield subset mini-network: pathogen, root disease, foliar disease, soil, and yield"
  ) +
  ggplot2::theme_void()

ggplot2::ggsave(
  "fig_yield_mini_network.png",
  p_yield_net,
  width = 9,
  height = 6,
  dpi = 300
)

```

# -----------------------------
# 4. PCA
# -----------------------------
```{r}
#-----------------------------
# Shared PCA input with manuscript-style names
#-----------------------------
yield_pca_df <- df %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na() %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "SCN density"                   = scn_count_rep,
    "Soil pH"                       = soil_ph,
    "FDX"                           = fdx,
    "RDX"                           = rdx,
    "Yield"                         = yield_kgha
  )

yield_pca_res <- FactoMineR::PCA(
  yield_pca_df,
  scale.unit = TRUE,
  graph = FALSE
)

yield_pca_loadings <- as.data.frame(yield_pca_res$var$coord) %>%
  tibble::rownames_to_column("variable")

readr::write_csv(yield_pca_loadings, "yield_pca_loadings.csv")

#-----------------------------
# Scree plot
#-----------------------------
p_yield_scree <- factoextra::fviz_eig(yield_pca_res, addlabels = TRUE) +
  ggplot2::labs(
    title = "Yield subset PCA scree plot",
    x = "Principal components",
    y = "Percentage of explained variance"
  ) +
  theme_siteyear

ggplot2::ggsave(
  "fig_yield_pca_scree.png",
  p_yield_scree,
  width = 7,
  height = 5,
  dpi = 300
)

#-----------------------------
# PCA biplot by FDX level
#-----------------------------
yield_pca_fdxlvl <- df_fdx_ph_mod %>%
  dplyr::select(
    fdx_level,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na() %>%
  dplyr::mutate(
    fdx_level = factor(
      fdx_level,
      levels = c("none", "low", "high"),
      labels = c("None (FDX = 0)", "Low (0 < FDX < 10)", "High (FDX ≥ 10)")
    )
  ) %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "SCN density"                   = scn_count_rep,
    "Soil pH"                       = soil_ph,
    "FDX"                           = fdx,
    "RDX"                           = rdx,
    "Yield"                         = yield_kgha
  )

yield_pca_res_fdxlvl <- FactoMineR::PCA(
  yield_pca_fdxlvl %>% dplyr::select(-fdx_level),
  scale.unit = TRUE,
  graph = FALSE
)

fdx_labels<-c("None (FDX = 0)", "Low (0 < FDX < 10)", "High (FDX ≥ 10)")
fdx_cols<-c( "#00BA38", "#B79F00", "#F8766D")

# Base plot without variable labels
p_base <- factoextra::fviz_pca_biplot(
  yield_pca_res_fdxlvl,
  geom.ind = "point",
  habillage = yield_pca_fdxlvl$fdx_level,
  addEllipses = TRUE,
  ellipse.alpha = 0.05,
  ellipse.level = 0.95,
  label = "none",
  col.var = "black",
  arrowsize = 0.5,
  pointshape = 16,
  pointsize = 1.25,
  alpha.ind = 0.65,
  palette = fdx_cols
) 

# Find the layer that contains the displayed arrow endpoints
gb <- ggplot_build(p_base)
arrow_layer_idx <- which(vapply(
  gb$data,
  function(d) all(c("x", "y", "xend", "yend") %in% names(d)),
  logical(1)
))[1]

arrow_df <- gb$data[[arrow_layer_idx]]

# Use the variable order from the PCA object
var_names <- rownames(yield_pca_res_fdxlvl$var$coord)

var_df <- dplyr::tibble(
  variable = var_names,
  x = arrow_df$xend * 1.04,
  y = arrow_df$yend * 1.04,
  hjust = dplyr::if_else(arrow_df$xend >= 0, 0, 1),
  vjust = dplyr::if_else(arrow_df$yend >= 0, 0, 1)
) %>%
  dplyr::mutate(
    lab = dplyr::case_when(
      variable == "F. virguliforme" ~ "F. virguliforme",
      TRUE ~ variable
    )
  )

var_df <- var_df %>%
  dplyr::mutate(
    x = dplyr::case_when(
      variable == "Soil pH" ~ x - 0.35,
      variable == "FDX"     ~ x - 0.20,
      variable == "RDX"     ~ x + 0.15,
      variable == "F. virguliforme" ~ x + 0.5,
      TRUE ~ x
    ),
    y = dplyr::case_when(
      variable == "Soil pH" ~ y + 0.10,
      variable == "FDX"     ~ y + 0.18,
      variable == "RDX"     ~ y - 0.08,
      TRUE ~ y
    )
  )

p_yield_pca_fdxlvl <- p_base +
  theme_siteyear +
  ggplot2::theme(
    legend.title = element_text(size = 11),
    legend.key.size = grid::unit(0.6, "cm")
  ) +
  scale_color_manual(
  values = fdx_cols,
  labels = fdx_labels,
  name = expression(atop("Foliar disease", "(FDX) level")),
  drop = FALSE
  ) +
  scale_fill_manual(
    values = fdx_cols,
    labels = fdx_labels,
    name = expression(atop("Foliar disease", "(FDX) level")),
    drop = FALSE,
    guide = "none"
  ) +
  coord_cartesian(xlim = c(-6.0, 6.0),ylim=c(-4.0,4.0)) +
  ggplot2::geom_text(
    data = var_df,
    aes(x = x, y = y, label = lab, hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  ) +
  ggplot2::labs(
    title = "PCA biplot by FDX level",
    x = paste0("PC1 (", round(yield_pca_res_fdxlvl$eig[1, 2], 1), "% variance)"),
    y = paste0("PC2 (", round(yield_pca_res_fdxlvl$eig[2, 2], 1), "% variance)")
  )

ggplot2::ggsave(
  "fig_yield_pca_biplot_fdx_level.png",
  p_yield_pca_fdxlvl,
  width = 8,
  height = 6,
  dpi = 300
)

#-----------------------------
# PCA biplot by SCN level
#-----------------------------
yield_pca_scn <- df %>%
  dplyr::select(
    scn_level,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na() %>%
  dplyr::filter(!is.na(scn_level)) %>%
  dplyr::mutate(
    scn_level = factor(
      scn_level,
      levels = c("none","low", "high"),
      labels = c("No SCN", "Low SCN", "High SCN")
    )
  ) %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "SCN density"                   = scn_count_rep,
    "Soil pH"                       = soil_ph,
    "FDX"                           = fdx,
    "RDX"                           = rdx,
    "Yield"                         = yield_kgha
  )

yield_pca_res_scn <- FactoMineR::PCA(
  yield_pca_scn %>% dplyr::select(-scn_level),
  scale.unit = TRUE,
  graph = FALSE
)

scn_labels <- c(
  "No SCN"   = "No SCN detected",
  "Low SCN" = expression("Low SCN (" * "eggs + J2∙100" * cm^{3} * ")"),
  "High SCN"  = expression("High SCN (" * "egs + J2∙100" * cm^{3} * ")")
)

scn_cols <- c(
  "No SCN"   = "#00BA38",   #
  "Low SCN" = "#B79F00",   # 
  "High SCN" = "#F8766D"   # 
)

p_yield_pca_scn <- factoextra::fviz_pca_biplot(
  yield_pca_res_scn,
  geom.ind = "point",
  habillage = yield_pca_scn$scn_level,
  addEllipses = TRUE,
  ellipse.alpha = 0.05,
  ellipse.level = 0.95,
  label = "none",
  col.var = "black",
  arrowsize = 0.5,
  pointshape = 16,
  pointsize = 1.25,
  alpha.ind = 0.65,
  palette = scn_cols
)  

# Base plot without variable labels
p_base <- factoextra::fviz_pca_biplot(
  yield_pca_res_scn,
  geom.ind = "point",
  habillage = yield_pca_scn$scn_level,
  addEllipses = TRUE,
  ellipse.alpha = 0.05,
  ellipse.level = 0.95,
  label = "none",
  col.var = "black",
  arrowsize = 0.5,
  pointshape = 16,
  pointsize = 1.25,
  alpha.ind = 0.65,
  palette = scn_cols
) 

# Find the layer that contains the displayed arrow endpoints
gb <- ggplot_build(p_base)
arrow_layer_idx <- which(vapply(
  gb$data,
  function(d) all(c("x", "y", "xend", "yend") %in% names(d)),
  logical(1)
))[1]

arrow_df <- gb$data[[arrow_layer_idx]]

# Use the variable order from the PCA object
var_names <- rownames(yield_pca_res_scn$var$coord)

var_df <- dplyr::tibble(
  variable = var_names,
  x = arrow_df$xend * 1.04,
  y = arrow_df$yend * 1.04,
  hjust = dplyr::if_else(arrow_df$xend >= 0, 0, 1),
  vjust = dplyr::if_else(arrow_df$yend >= 0, 0, 1)
) %>%
  dplyr::mutate(
    lab = dplyr::case_when(
      variable == "F. virguliforme" ~ "F. virguliforme",
      TRUE ~ variable
    )
  )

var_df <- var_df %>%
  dplyr::mutate(
    x = dplyr::case_when(
      variable == "Soil pH" ~ x - 0.35,
      variable == "FDX"     ~ x - 0.20,
      variable == "RDX"     ~ x + 0.15,
      variable == "F. virguliforme" ~ x + 0.5,
      TRUE ~ x
    ),
    y = dplyr::case_when(
      variable == "Soil pH" ~ y + 0.10,
      variable == "FDX"     ~ y + 0.18,
      variable == "RDX"     ~ y - 0.08,
      TRUE ~ y
    )
  )

p_yield_pca_scn <- p_base +
  theme_siteyear +
  ggplot2::theme(
    legend.title = element_text(size = 11),
    legend.key.size = grid::unit(0.6, "cm")
  ) +
  scale_color_manual(
  values = scn_cols,
  labels = scn_labels,
  name = expression(atop(italic("Heterodera glycines"),
                           "(SCN) density level")),
  drop = FALSE
  ) +
  scale_fill_manual(
    values = scn_cols,
    labels = scn_labels,
    name = expression(atop(italic("Heterodera glycines"),
                           "(SCN) density level")),
    drop = FALSE,
    guide = "none"
  ) +
  coord_cartesian(xlim = c(-6.0, 6.0),ylim=c(-4.0,4.0)) +
  ggplot2::geom_text(
    data = var_df,
    aes(x = x, y = y, label = lab, hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  ) +
  ggplot2::labs(
    title = "PCA biplot by SCN level",
    x = paste0("PC1 (", round(yield_pca_res_scn$eig[1, 2], 1), "% variance)"),
    y = paste0("PC2 (", round(yield_pca_res_scn$eig[2, 2], 1), "% variance)")
  )

ggplot2::ggsave(
  "fig_yield_pca_biplot_scn_level.png",
  p_yield_pca_scn,
  width = 8,
  height = 6,
  dpi = 300
)

#-----------------------------
# PCA biplot by simplified Fv class
# Strict numeric alignment:
# 0, 0-1, 1-3, 3-5, >5
#-----------------------------
yield_pca_fv_simple <- df_fdx_ph_mod %>%
  dplyr::select(
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na() %>%
  dplyr::mutate(
    fv_level_simple = dplyr::case_when(
      ratio_fvfg_dnang == 0 ~ "0",
      ratio_fvfg_dnang > 0 & ratio_fvfg_dnang < 1 ~ "0-1",
      ratio_fvfg_dnang >= 1 & ratio_fvfg_dnang < 3 ~ "1-3",
      ratio_fvfg_dnang >= 3 & ratio_fvfg_dnang < 5 ~ "3-5",
      ratio_fvfg_dnang >= 5 ~ ">5"
    ),
    fv_level_simple = factor(
      fv_level_simple,
      levels = c("0", "0-1", "1-3", "3-5", ">5")
    )
  ) %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "SCN density"     = scn_count_rep,
    "Soil pH"         = soil_ph,
    "FDX"             = fdx,
    "RDX"             = rdx,
    "Yield"           = yield_kgha
  )

yield_pca_res_fv_simple <- FactoMineR::PCA(
  yield_pca_fv_simple %>% dplyr::select(-fv_level_simple),
  scale.unit = TRUE,
  graph = FALSE
)

fv_labels <- c(
  "0"   = "0 (undetected)",
  "0-1" = expression("Low (" * "0–1" ~ fg %.% ng^{-1} * ")"),
  "1-3" = expression("Moderate (" * "1–3" ~ fg %.% ng^{-1} * ")"),
  "3-5" = expression("Moderate-High (" * "3–5" ~ fg %.% ng^{-1} * ")"),
  ">5"  = expression("High (" * ">" * 5 ~ fg %.% ng^{-1} * ")")
)


fv_cols <- c(
  "0"   = "#4169E1",   # royal blue
  "0-1" = "#00A651",   # darker green
  "1-3" = "#C49A25",   # deeper gold
  "3-5" = "#E26B20",   # red-orange
  ">5"  = "#850101"    # deep red
)

# Base plot without variable labels
p_base <- factoextra::fviz_pca_biplot(
  yield_pca_res_fv_simple,
  geom.ind = "point",
  habillage = yield_pca_fv_simple$fv_level_simple,
  addEllipses = TRUE,
  ellipse.alpha = 0.05,
  ellipse.level = 0.95,
  label = "none",
  col.var = "black",
  arrowsize = 0.5,
  pointshape = 16,
  pointsize = 1.25,
  alpha.ind = 0.65,
  palette = fv_cols
)

# Find the layer that contains the displayed arrow endpoints
gb <- ggplot_build(p_base)
arrow_layer_idx <- which(vapply(
  gb$data,
  function(d) all(c("x", "y", "xend", "yend") %in% names(d)),
  logical(1)
))[1]

arrow_df <- gb$data[[arrow_layer_idx]]

# Use the variable order from the PCA object
var_names <- rownames(yield_pca_res_fv_simple$var$coord)

var_df <- dplyr::tibble(
  variable = var_names,
  x = arrow_df$xend * 1.04,
  y = arrow_df$yend * 1.04,
  hjust = dplyr::if_else(arrow_df$xend >= 0, 0, 1),
  vjust = dplyr::if_else(arrow_df$yend >= 0, 0, 1)
) %>%
  dplyr::mutate(
    lab = dplyr::case_when(
      variable == "F. virguliforme" ~ "F. virguliforme",
      TRUE ~ variable
    )
  )

var_df <- var_df %>%
  dplyr::mutate(
    x = dplyr::case_when(
      variable == "Soil pH" ~ x - 0.35,
      variable == "FDX"     ~ x - 0.20,
      variable == "RDX"     ~ x + 0.15,
      variable == "F. virguliforme" ~ x + 0.5,
      TRUE ~ x
    ),
    y = dplyr::case_when(
      variable == "Soil pH" ~ y + 0.10,
      variable == "FDX"     ~ y + 0.18,
      variable == "RDX"     ~ y - 0.08,
      TRUE ~ y
    )
  )

p_yield_pca_fv_simple <- p_base +
  theme_siteyear +
  ggplot2::theme(
    legend.title = element_text(size = 11),
    legend.key.size = grid::unit(0.6, "cm")
  ) +
  scale_color_manual(
    values = fv_cols,
    labels = fv_labels,
    name = expression(atop(italic("Fusarium virguliforme"),
                           "DNA abundance level")),
    drop = FALSE
  ) +
  scale_fill_manual(
    values = fv_cols,
    labels = fv_labels,
    name = expression(atop(italic("F. virguliforme"),
                           "DNA abundance level")),
    drop = FALSE,
    guide = "none"
  ) +
  coord_cartesian(xlim = c(-6.0, 6.0),ylim=c(-4.0,4.0)) +
  ggplot2::geom_text(
    data = var_df,
    aes(x = x, y = y, label = lab, hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  ) +
  ggplot2::labs(
    title = expression(paste(
    "PCA biplot by ", italic("Fusarium virguliforme"), " DNA abundance level"
  )),
    x = paste0("PC1 (", round(yield_pca_res_fv_simple$eig[1, 2], 1), "% variance)"),
    y = paste0("PC2 (", round(yield_pca_res_fv_simple$eig[2, 2], 1), "% variance)")
  )

ggplot2::ggsave(
  "fig_yield_pca_biplot_fv_simple.png",
  p_yield_pca_fv_simple,
  width = 8,
  height = 6,
  dpi = 600
)

#-----------------------------
# PCA biplot by product
#-----------------------------
yield_pca_prod <- df %>%
  dplyr::select(
    product,
    ratio_fvfg_dnang,
    scn_count_rep,
    soil_ph,
    fdx,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na() %>%
  dplyr::mutate(
    product = factor(
      product,
      levels = c("Base", "ILEVO", "Saltro"),
      labels = c("CB", "CB + fluopyram", "CB + pydiflumetofen")
    )
  ) %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "SCN density"                   = scn_count_rep,
    "Soil pH"                       = soil_ph,
    "FDX"                           = fdx,
    "RDX"                           = rdx,
    "Yield"                         = yield_kgha
  )

yield_pca_res_prod <- FactoMineR::PCA(
  yield_pca_prod %>% dplyr::select(-product),
  scale.unit = TRUE,
  graph = FALSE
)

trt_cols<-c("#F8766D", "#00BA38", "#619CFF")
trt_labels<-c("CB", "CB + fluopyram", "CB + pydiflumetofen")

p_yield_pca_prod <- factoextra::fviz_pca_biplot(
  yield_pca_res_prod,
  geom.ind = "point",
  habillage = yield_pca_prod$product,
  addEllipses = TRUE,
  ellipse.level = 0.95,
  label = "var",
  repel = TRUE,
  col.var = "black",
  arrowsize = 0.7,
  pointshape = 16,
  pointsize = 2.2,
  alpha.ind = 0.65,
  palette = trt_cols
) 

# Base plot without variable labels
p_base <- factoextra::fviz_pca_biplot(
  yield_pca_res_prod,
  geom.ind = "point",
  habillage = yield_pca_prod$product,
  addEllipses = TRUE,
  ellipse.alpha = 0.05,
  ellipse.level = 0.95,
  label = "none",
  col.var = "black",
  arrowsize = 0.5,
  pointshape = 16,
  pointsize = 1.25,
  alpha.ind = 0.65,
  palette = trt_cols
) 

# Find the layer that contains the displayed arrow endpoints
gb <- ggplot_build(p_base)
arrow_layer_idx <- which(vapply(
  gb$data,
  function(d) all(c("x", "y", "xend", "yend") %in% names(d)),
  logical(1)
))[1]

arrow_df <- gb$data[[arrow_layer_idx]]

# Use the variable order from the PCA object
var_names <- rownames(yield_pca_res_prod$var$coord)

var_df <- dplyr::tibble(
  variable = var_names,
  x = arrow_df$xend * 1.04,
  y = arrow_df$yend * 1.04,
  hjust = dplyr::if_else(arrow_df$xend >= 0, 0, 1),
  vjust = dplyr::if_else(arrow_df$yend >= 0, 0, 1)
) %>%
  dplyr::mutate(
    lab = dplyr::case_when(
      variable == "F. virguliforme" ~ "F. virguliforme",
      TRUE ~ variable
    )
  )

var_df <- var_df %>%
  dplyr::mutate(
    x = dplyr::case_when(
      variable == "Soil pH" ~ x - 0.35,
      variable == "FDX"     ~ x - 0.20,
      variable == "RDX"     ~ x + 0.15,
      variable == "F. virguliforme" ~ x + 0.5,
      TRUE ~ x
    ),
    y = dplyr::case_when(
      variable == "Soil pH" ~ y + 0.10,
      variable == "FDX"     ~ y + 0.18,
      variable == "RDX"     ~ y - 0.08,
      TRUE ~ y
    )
  )

p_yield_pca_prod <- p_base +
  theme_siteyear +
  ggplot2::theme(
    legend.title = element_text(size = 11),
    legend.key.size = grid::unit(0.6, "cm")
  ) +
  scale_color_manual(
  values = trt_cols,
  labels = trt_labels,
  name = c("Seed treatment"),
  drop = FALSE
  ) +
  scale_fill_manual(
    values = trt_cols,
    labels = trt_labels,
    name = c("Seed treatment"),
    drop = FALSE,
    guide = "none"
  ) +
  coord_cartesian(xlim = c(-6.0, 6.0),ylim=c(-4.0,4.0)) +
  ggplot2::geom_text(
    data = var_df,
    aes(x = x, y = y, label = lab, hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    size = 4,
    color = "black"
  ) +
    ggplot2::labs(
    title = "Yield subset PCA biplot by seed treatment product",
    x = paste0("PC1 (", round(yield_pca_res_prod$eig[1, 2], 1), "% variance)"),
    y = paste0("PC2 (", round(yield_pca_res_prod$eig[2, 2], 1), "% variance)")
  )

ggplot2::ggsave(
  "fig_yield_pca_biplot_product.png",
  p_yield_pca_prod,
  width = 8,
  height = 6,
  dpi = 300
)

#-----------------------------
# PCA loadings bar plot for first 4 PCs
#-----------------------------
pc_var <- yield_pca_res_fv_simple$eig[1:4, 2]  # % variance
pc_labels <- paste0(
  "PC", 1:4, " (", sprintf("%.1f", pc_var), "%)"
)

# Extract loadings
yield_pca_loadings_plot <- as.data.frame(yield_pca_res_fv_simple$var$coord) %>%
  tibble::rownames_to_column("variable") %>%
  dplyr::filter(
    !is.na(variable),
    variable != "",
    variable != "NA"
  ) %>%
  dplyr::select(variable, Dim.1, Dim.2, Dim.3, Dim.4) %>%
  dplyr::rename(
    PC1 = Dim.1,
    PC2 = Dim.2,
    PC3 = Dim.3,
    PC4 = Dim.4
  ) %>%
  tidyr::pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "loading"
  ) %>%
  dplyr::filter(!is.na(loading)) %>%
  dplyr::mutate(
    PC = factor(
      PC,
      levels = c("PC1", "PC2", "PC3", "PC4"),
      labels = pc_labels
    ),
    variable = factor(
      variable,
      levels = c("F. virguliforme", "SCN density", "Soil pH", "FDX", "RDX", "Yield")
    ),
    loading_sign = dplyr::if_else(loading < 0, "Negative", "Positive")
  ) %>%
  droplevels()

unique(as.data.frame(yield_pca_res_fv_simple$var$coord) %>%
  tibble::rownames_to_column("variable") %>%
  dplyr::pull(variable))

table(as.data.frame(yield_pca_res_fv_simple$var$coord) %>%
  tibble::rownames_to_column("variable") %>%
  dplyr::pull(variable), useNA = "ifany")

yield_pca_loadings_plot <- yield_pca_loadings_plot %>%
  dplyr::filter(!is.na(variable) & as.character(variable) != "NA")

# Plot
p_yield_pca_loadings <- ggplot(
  yield_pca_loadings_plot,
  aes(x = loading, y = forcats::fct_rev(variable), fill = loading)
) +
  geom_col(width = 0.75, color = "black", linewidth = 0.25) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4) +
  facet_wrap(~ PC, ncol = 2, scales = "fixed") +
  scale_fill_gradient2(
    low = "#B2182B",   # red = negative
    mid = "white",
    high = "#2166AC",  # blue = positive
    midpoint = 0,
    name = "Loading"
  ) +
  geom_text(
  aes(label = sprintf(" %.2f", loading),
      hjust = if_else(loading < 0, 1.05, -0.05)),
  size = 3
  ) +
  coord_cartesian(xlim = c(-1.1, 1.1)) +
  labs(
    title = "PCA variable loadings for the first four principal components",
    x = "Loading",
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.2),
    plot.title = element_text(face = "bold", size = 12),
    strip.text = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "right"
  )

p_yield_pca_loadings

ggsave(
  "fig_yield_pca_loadings_barplot.png",
  p_yield_pca_loadings,
  width = 8,
  height = 6,
  dpi = 300
)

library(patchwork)

p_sup <- (
  p_yield_pca_fdxlvl +
    ggtitle("A") +
    theme(plot.title = element_text(size = 16, hjust = 0, face = "bold"))
) +
(
  p_yield_pca_scn +
    ggtitle("B") +
    theme(plot.title = element_text(size = 16, hjust = 0, face = "bold"))
) +
(
  p_yield_pca_prod +
    ggtitle("C") +
    theme(plot.title = element_text(size = 16, hjust = 0, face = "bold"))
) +
(
  p_yield_pca_loadings +
    ggtitle("D") +
    theme(plot.title = element_text(size = 16, hjust = 0, face = "bold"))
) +
plot_layout(ncol = 2)

ggsave(
  "fig_pca_supplemental.png",
  p_sup,
  width = 14,
  height = 8,
  dpi = 300
)

eig <- yield_pca_res_fv_simple$eig

pc3_var <- eig[3, "percentage of variance"]
pc4_var <- eig[4, "percentage of variance"]
pc34_sum <- pc3_var + pc4_var

loadings <- as.data.frame(yield_pca_res_fv_simple$var$coord)

loadings$variable <- rownames(loadings)

loadings %>%
  select(variable, Dim.1, Dim.2, Dim.3, Dim.4)

loadings %>%
  arrange(desc(abs(Dim.3))) %>% head()


eig <- yield_pca_res_fv_simple$eig

pc3 <- eig[3, "percentage of variance"]
pc4 <- eig[4, "percentage of variance"]
pc34 <- pc3 + pc4

pc3; pc4; pc34


```



#### corr matrices ####
```{r}
#========================================================
# Final polished publication chunk:
# heterogeneous core matrix + full mixed-type matrix
# with validation pairs, no identity diagonal,
# manual diagonal labels for clean placement
#========================================================

#--------------------------------------------------------
# Options / toggles
#--------------------------------------------------------
SHOW_TYPE_CODES_CORE <- TRUE
SHOW_TYPE_CODES_FULL <- TRUE
HIGHLIGHT_VALIDATION_PAIRS <- TRUE

#--------------------------------------------------------
# Helper functions
#--------------------------------------------------------
is_binary_factor <- function(x) is.factor(x) && nlevels(x) == 2
is_ordinal       <- function(x) is.ordered(x)
is_unordered     <- function(x) is.factor(x) && !is_ordinal(x)

eta_with_p <- function(num, fac) {
  fac <- droplevels(fac)
  ok <- !is.na(num) & !is.na(fac)
  num <- num[ok]
  fac <- fac[ok]

  if (length(num) < 2 || nlevels(fac) < 2) {
    return(list(stat = NA_real_, p = NA_real_))
  }

  fit <- aov(num ~ fac)
  tb  <- summary(fit)[[1]]
  SS_g <- tb["fac", "Sum Sq"]
  SS_e <- tb["Residuals", "Sum Sq"]
  eta2p <- SS_g / (SS_g + SS_e)

  list(
    stat = sqrt(max(0, eta2p)),
    p    = tb["fac", "Pr(>F)"]
  )
}

cramer_v <- function(x, y) {
  tab <- table(x, y)
  chi <- suppressWarnings(chisq.test(tab, correct = FALSE))
  n   <- sum(tab)
  r   <- nrow(tab)
  c   <- ncol(tab)
  V   <- sqrt(as.numeric(chi$statistic) / (n * (min(r, c) - 1)))
  list(stat = V, p = chi$p.value)
}

poly_serial_safe <- function(x, ord) {
  if (requireNamespace("polycor", quietly = TRUE)) {
    out <- try(polycor::polyserial(x, as.ordered(ord)), silent = TRUE)
    if (!inherits(out, "try-error")) {
      return(list(stat = as.numeric(out), p = NA_real_))
    }
  }

  z  <- suppressWarnings(as.numeric(ord))
  ct <- suppressWarnings(cor.test(x, z))
  list(stat = unname(ct$estimate), p = unname(ct$p.value))
}

poly_choric_safe <- function(x, y) {
  if (requireNamespace("polycor", quietly = TRUE)) {
    out <- try(
      polycor::polychor(as.ordered(x), as.ordered(y), ML = TRUE, std.err = TRUE),
      silent = TRUE
    )
    if (!inherits(out, "try-error") && !is.null(attr(out, "std.err"))) {
      rho <- as.numeric(out)
      se  <- attr(out, "std.err")
      z   <- rho / se
      p   <- 2 * pnorm(-abs(z))
      return(list(stat = rho, p = p))
    }
  }

  cramer_v(as.factor(x), as.factor(y))
}

assoc_pair <- function(x, y) {

  if (is.numeric(x) && is.numeric(y)) {
    ct <- suppressWarnings(cor.test(x, y, method = "spearman", exact = FALSE))
    return(list(stat = unname(ct$estimate), p = unname(ct$p.value), type = "r"))
  }

  if (is.numeric(x) && is_binary_factor(y)) {
    b  <- as.numeric(y) - 1
    ct <- suppressWarnings(cor.test(x, b))
    return(list(stat = unname(ct$estimate), p = unname(ct$p.value), type = "pb"))
  }

  if (is.numeric(y) && is_binary_factor(x)) {
    b  <- as.numeric(x) - 1
    ct <- suppressWarnings(cor.test(y, b))
    return(list(stat = unname(ct$estimate), p = unname(ct$p.value), type = "pb"))
  }

  if (is.numeric(x) && is_unordered(y) && nlevels(y) > 2) {
    res <- eta_with_p(x, y)
    return(list(stat = res$stat, p = res$p, type = "η"))
  }

  if (is.numeric(y) && is_unordered(x) && nlevels(x) > 2) {
    res <- eta_with_p(y, x)
    return(list(stat = res$stat, p = res$p, type = "η"))
  }

  if (is.numeric(x) && is_ordinal(y)) {
    res <- poly_serial_safe(x, y)
    return(list(stat = res$stat, p = res$p, type = "ps"))
  }

  if (is.numeric(y) && is_ordinal(x)) {
    res <- poly_serial_safe(y, x)
    return(list(stat = res$stat, p = res$p, type = "ps"))
  }

  if (is_ordinal(x) && is_ordinal(y)) {
    res <- poly_choric_safe(x, y)
    return(list(stat = res$stat, p = res$p, type = "pc"))
  }

  res <- cramer_v(as.factor(x), as.factor(y))
  list(stat = res$stat, p = res$p, type = "V")
}

#--------------------------------------------------------
# Build correlation-ready dataframe with stratification vars
#--------------------------------------------------------
df_corr <- df %>%
  dplyr::mutate(
    fv_level = dplyr::case_when(
      ratio_fvfg_dnang == 0 ~ "0",
      ratio_fvfg_dnang > 0 & ratio_fvfg_dnang < 1 ~ "0–1",
      ratio_fvfg_dnang >= 1 & ratio_fvfg_dnang < 3 ~ "1–3",
      ratio_fvfg_dnang >= 3 & ratio_fvfg_dnang < 5 ~ "3–5",
      ratio_fvfg_dnang >= 5 ~ ">5"
    ),
    fv_level = factor(fv_level, levels = c("0", "0–1", "1–3", "3–5", ">5")),

    scn_level = dplyr::case_when(
      scn_count_rep == 0 ~ "None",
      scn_count_rep > 0 & scn_count_rep < 2000 ~ "Low",
      scn_count_rep >= 2000 ~ "High"
    ),
    scn_level = factor(scn_level, levels = c("None", "Low", "High")),

    fdx_level = dplyr::case_when(
      fdx == 0 ~ "None",
      fdx > 0 & fdx < 10 ~ "Low",
      fdx >= 10 ~ "High"
    ),
    fdx_level = factor(fdx_level, levels = c("None", "Low", "High")),

    sds_tol = factor(sds_tol),
    product = factor(product)
  )

#--------------------------------------------------------
# Shared helper to build pairwise association dataframe
#--------------------------------------------------------
build_assoc_df <- function(dat, var_names) {
  expand.grid(
    var1 = var_names,
    var2 = var_names,
    stringsAsFactors = FALSE
  ) %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      res = list(assoc_pair(dat[[var1]], dat[[var2]])),
      coef = res$stat,
      p    = res$p,
      type = res$type
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      sig = dplyr::case_when(
        !is.na(p) & p < 0.001 ~ "***",
        !is.na(p) & p < 0.01  ~ "**",
        !is.na(p) & p < 0.05  ~ "*",
        TRUE                  ~ ""
      ),
      coef_lab = sprintf("%.2f", coef),
      type_lab = dplyr::case_when(
        type %in% c("r", "pb", "η", "ps", "pc", "V") ~ type,
        TRUE ~ ""
      )
    )
}

#--------------------------------------------------------
# Shared helper for validation pair flagging
#--------------------------------------------------------
flag_validation_pairs <- function(df_assoc) {
  df_assoc %>%
    dplyr::mutate(
      validation_pair = dplyr::case_when(
        (var1 == "F. virguliforme" & var2 == "Fv level") |
          (var2 == "F. virguliforme" & var1 == "Fv level") ~ TRUE,

        (var1 == "SCN density" & var2 == "SCN level") |
          (var2 == "SCN density" & var1 == "SCN level") ~ TRUE,

        (var1 == "FDX" & var2 == "FDX level") |
          (var2 == "FDX" & var1 == "FDX level") ~ TRUE,

        TRUE ~ FALSE
      )
    )
}

#--------------------------------------------------------
# Shared plotting helper
#--------------------------------------------------------
plot_assoc_matrix_gg <- function(df_assoc,
                                 var_names,
                                 title_text,
                                 fill_title = "Association",
                                 show_type_codes = TRUE,
                                 highlight_validation_pairs = TRUE,
                                 x_limits = NULL,
                                 y_limits = NULL,
                                 value_size = 3.8,
                                 type_size  = 2.5,
                                 star_size  = 3.2) {

  n_vars <- length(var_names)

  df_plot <- df_assoc %>%
    dplyr::mutate(
      row_id = match(var1, var_names),
      col_id = match(var2, var_names),
      x = col_id,
      y = n_vars - row_id + 1
    ) %>%
    dplyr::mutate(
      keep_cell =
        (row_id < col_id) |
        (validation_pair & row_id == col_id - 1)
    ) %>%
    dplyr::filter(keep_cell)

  if (is.null(x_limits)) x_limits <- c(-0.8, n_vars + 0.9)
  if (is.null(y_limits)) y_limits <- c(0.2, n_vars + 1.15)

  p <- ggplot() +
    geom_tile(
      data = df_plot,
      aes(
        x = x,
        y = y,
        fill = coef,
        color = if (highlight_validation_pairs) validation_pair else FALSE
      ),
      linewidth = if (highlight_validation_pairs) 0.95 else 0.5,
      width = 0.98,
      height = 0.98
    ) +
    geom_text(
      data = df_plot,
      aes(x = x, y = y, label = coef_lab),
      size = value_size,
      fontface = "bold"
    ) +
    geom_text(
      data = df_plot,
      aes(x = x, y = y, label = sig),
      nudge_y = -0.28,
      size = star_size
    ) +
    scale_fill_gradient2(
      low = "#B2182B",
      mid = "white",
      high = "#2166AC",
      midpoint = 0,
      limits = c(-1, 1),
      name = fill_title
    ) +
    scale_color_manual(
      values = c("TRUE" = "black", "FALSE" = "white"),
      guide = "none"
    ) +
    scale_x_continuous(
      limits = x_limits,
      breaks = NULL,
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      limits = y_limits,
      breaks = NULL,
      expand = c(0, 0)
    ) +
    labs(
      title = title_text,
      x = NULL,
      y = NULL
    ) +
    coord_fixed(clip = "off") +
    theme_bw(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      panel.border = element_blank(),
      plot.title = element_text(face = "bold", size = 12, margin = margin(0, 0, 45, 0)),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.title = element_text(face = "bold"),
      plot.margin = margin(10, 30, 10, 30),
      legend.margin = margin(0, 0, 0, 0)
        )

  if (show_type_codes) {
    p <- p +
      geom_text(
        data = df_plot,
        aes(x = x, y = y, label = type_lab),
        nudge_x = -0.36,
        nudge_y = 0.30,
        hjust = 0,
        vjust = 1,
        size = type_size,
        color = "black"
      )
  }

  p
}


#========================================================
# 1. CORE MATRIX
#========================================================
corr_vars_core <- df_corr %>%
  dplyr::select(
    ratio_fvfg_dnang,
    fv_level,
    scn_count_rep,
    scn_level,
    soil_ph,
    fdx,
    fdx_level,
    rdx,
    yield_kgha
  ) %>%
  tidyr::drop_na() %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "Fv level"        = fv_level,
    "SCN density"     = scn_count_rep,
    "SCN level"       = scn_level,
    "Soil pH"         = soil_ph,
    "FDX"             = fdx,
    "FDX level"       = fdx_level,
    "RDX"             = rdx,
    "Yield"           = yield_kgha
  )

var_names_core <- names(corr_vars_core)

corr_df_core <- build_assoc_df(corr_vars_core, var_names_core) %>%
  flag_validation_pairs()

top_labels_core <- tibble::tibble(
  label = var_names_core[-1],                  # drop F. virguliforme
  x = seq_along(var_names_core)[-1],
  y = length(var_names_core) + 0.78
)

core_diag_labels <- tibble::tribble(
  ~label,              ~x, ~y,  ~hjust, ~vjust,
  "F. virguliforme",    0,  9,   0.25,   0.25,
  "Fv level",           2,  8,   0.50,   0.50,
  "SCN density",        3,  7,   0.75,   0.50,
  "SCN level",          4,  6,   0.75,   0.50,
  "Soil pH",            5,  5,   0.50,   0.50,
  "FDX",                6,  4,   0.50,   0.50,
  "FDX level",          7,  3,   0.75,   0.50,
  "RDX",                8,  2,   0.50,   0.50,
  #"Yield",              9,  1,   0.50,   0.50
)

p_corr_core <- plot_assoc_matrix_gg(
  df_assoc = corr_df_core,
  var_names = var_names_core,
  title_text = "Heterogeneous association matrix of core biological variables",
  fill_title = "Association",
  show_type_codes = SHOW_TYPE_CODES_CORE,
  highlight_validation_pairs = HIGHLIGHT_VALIDATION_PAIRS,
  x_limits = c(-1.10, length(var_names_core) + 0.95),
  y_limits = c(0.2, length(var_names_core) + 1.10),
  value_size = 3.8,
  type_size  = 2.5,
  star_size  = 3.1
) +
  geom_text(
    data = core_diag_labels,
    aes(x = x, y = y, label = label, hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    size = 3.0,
    color = "black"
  ) +
  geom_text(
    data = top_labels_core,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    angle = 45,
    hjust = 0,
    vjust = 0,
    size = 3.0
  )

ggsave(
  "fig_corr_matrix_core_validation.png",
  p_corr_core,
  width = 8.8,
  height = 7.5,
  dpi = 300
)

#========================================================
# 2. FULL MATRIX
#========================================================
corr_vars_full <- df_corr %>%
  dplyr::select(
    ratio_fvfg_dnang,
    fv_level,
    scn_count_rep,
    scn_level,
    soil_ph,
    fdx,
    fdx_level,
    rdx,
    yield_kgha,
    sds_tol,
    product
  ) %>%
  tidyr::drop_na() %>%
  dplyr::rename(
    "F. virguliforme" = ratio_fvfg_dnang,
    "Fv level"        = fv_level,
    "SCN density"     = scn_count_rep,
    "SCN level"       = scn_level,
    "Soil pH"         = soil_ph,
    "FDX"             = fdx,
    "FDX level"       = fdx_level,
    "RDX"             = rdx,
    "Yield"           = yield_kgha,
    "SDS tolerance"   = sds_tol,
    "Seed treatment"  = product
  )

var_names_full <- names(corr_vars_full)

corr_df_full <- build_assoc_df(corr_vars_full, var_names_full) %>%
  flag_validation_pairs()

top_labels_full <- tibble::tibble(
  label = var_names_full[-1],                  # drop F. virguliforme
  x = seq_along(var_names_full)[-1],
  y = length(var_names_full) + 0.78
)

full_diag_labels <- tibble::tribble(
  ~label,              ~x, ~y,  ~hjust, ~vjust,
  "F. virguliforme",    0, 11,   0.10,   0.25,
  "Fv level",           2, 10,   0.50,   0.50,
  "SCN density",        3,  9,   0.75,   0.50,
  "SCN level",          4,  8,   0.75,   0.50,
  "Soil pH",            5,  7,   0.50,   0.50,
  "FDX",                6,  6,   0.50,   0.50,
  "FDX level",          7,  5,   0.75,   0.50,
  "RDX",                8,  4,   0.50,   0.50,
  "Yield",              9,  3,   0.50,   0.50,
  "SDS tolerance",     10,  2,   0.75,   0.50
  # Seed treatment intentionally omitted from diagonal labels
)

p_corr_full <- plot_assoc_matrix_gg(
  df_assoc = corr_df_full,
  var_names = var_names_full,
  title_text = "Mixed-type association matrix with validation pairs",
  fill_title = "Association",
  show_type_codes = SHOW_TYPE_CODES_FULL,
  highlight_validation_pairs = HIGHLIGHT_VALIDATION_PAIRS,
  x_limits = c(-1.30, length(var_names_full) + 1.10),
  y_limits = c(0.2, length(var_names_full) + 1.15),
  value_size = 3.8,
  type_size  = 2.5,
  star_size  = 3.1
) +
  geom_text(
    data = full_diag_labels,
    aes(x = x, y = y, label = label, hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    size = 3.0,
    color = "black"
  ) +
  geom_text(
    data = top_labels_full,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    angle = 45,
    hjust = 0,
    vjust = 0,
    size = 3.0
  )

ggsave(
  "fig_corr_matrix_full_validation.png",
  p_corr_full,
  width = 11.6,
  height = 9.8,
  dpi = 300
)


```


###surface response stuff###
```{r}

library(plotly)

surf_df <- df %>%
  mutate(
    fv_log = log10(ratio_fvfg_dnang + 1),
    scn_log = log10(scn_count_rep + 1),
    rdx_log1p = log1p(rdx),
    fdx_log1p = log1p(fdx)
  )

surf_rdx <- surf_df %>%
  filter(!is.na(fv_log), !is.na(scn_log), !is.na(rdx_log1p))

surf_fdx <- surf_df %>%
  filter(!is.na(fv_log), !is.na(scn_log), !is.na(fdx_log1p))

surf_yld <- surf_df %>%
  filter(!is.na(fv_log), !is.na(scn_log), !is.na(yield_kgha))

mod_rdx <- gam(rdx_log1p ~ s(fv_log, scn_log, k = 20), data = surf_rdx, method = "REML")
mod_fdx <- gam(fdx_log1p ~ s(fv_log, scn_log, k = 20), data = surf_fdx, method = "REML")
mod_yld <- gam(yield_kgha ~ s(fv_log, scn_log, k = 20), data = surf_yld, method = "REML")

fv_range  <- range(surf_df$fv_log, na.rm = TRUE)
scn_range <- range(surf_df$scn_log, na.rm = TRUE)

pred_grid <- expand.grid(
  fv_log  = seq(fv_range[1], fv_range[2], length.out = 100),
  scn_log = seq(scn_range[1], scn_range[2], length.out = 100)
)

pred_rdx <- pred_grid %>%
  mutate(
    fit = predict(mod_rdx, newdata = pred_grid, type = "response"),
    response = exp(fit) - 1,
    panel = "RDX"
  )

pred_fdx <- pred_grid %>%
  mutate(
    fit = predict(mod_fdx, newdata = pred_grid, type = "response"),
    response = exp(fit) - 1,
    panel = "FDX"
  )

pred_yld <- pred_grid %>%
  mutate(
    response = predict(mod_yld, newdata = pred_grid, type = "response"),
    panel = "Yield (kg ha^-1)"
  )

surface_panel <- function(pred_dat, obs_dat, zlab) {
  ggplot(pred_dat, aes(x = fv_log, y = scn_log, z = response)) +
    geom_raster(aes(fill = response), interpolate = TRUE) +
    geom_contour(color = "black", alpha = 0.6, bins = 8, linewidth = 0.3) +
    geom_point(
      data = obs_dat,
      aes(x = fv_log, y = scn_log),
      inherit.aes = FALSE,
      shape = 21, size = 1.2, stroke = 0.2,
      fill = alpha("white", 0.5), color = alpha("black", 0.5)
    ) +
    labs(
      x = expression(log[10]~"(F. virguliforme ratio + 1)"),
      y = expression(log[10]~"(SCN eggs + J2 per 100 cm"^3*" + 1)"),
      fill = zlab
    ) +
    theme_bw(base_size = 11) +
    theme(
      legend.position = "right",
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold")
    )
}

p_rdx <- surface_panel(pred_rdx, surf_rdx, "RDX") +
  ggtitle("A. Root disease index (RDX)")

p_fdx <- surface_panel(pred_fdx, surf_fdx, "FDX") +
  ggtitle("B. Foliar disease index (FDX)")

p_yld <- surface_panel(pred_yld, surf_yld, expression("Yield (kg ha"^-1*")")) +
  ggtitle("C. Yield")

surface_fig <- p_rdx / p_fdx / p_yld +
  plot_layout(ncol = 1, guides = "collect")

surface_fig

ggsave(
  filename = "response_surface_column_figure.png",
  plot = surface_fig,
  width = 8,
  height = 14,
  dpi = 600
)


# helper to convert long prediction grid to x/y/z for plotly surface
make_surface_matrix <- function(pred_dat) {
  x_vals <- sort(unique(pred_dat$fv_log))
  y_vals <- sort(unique(pred_dat$scn_log))
  
  z_mat <- pred_dat %>%
    arrange(scn_log, fv_log) %>%
    select(fv_log, scn_log, response) %>%
    pivot_wider(names_from = fv_log, values_from = response) %>%
    select(-scn_log) %>%
    as.matrix()
  
  list(x = x_vals, y = y_vals, z = z_mat)
}

surfmat_rdx <- make_surface_matrix(pred_rdx)
surfmat_fdx <- make_surface_matrix(pred_fdx)
surfmat_yld <- make_surface_matrix(pred_yld)

summary(pred_yld$response)
dim(surfmat_yld$z)
range(surfmat_yld$z, na.rm = TRUE)

```


### soil table for supps ###
```{r}

#========================================================
# Supplemental table: joint tests for final selected models
# Includes model name for traceability
#========================================================

#--------------------------------------------------------
# Helper: extract all fixed-effect joint tests from a model
#--------------------------------------------------------
extract_joint_tests <- function(model, response_label, model_name) {
  jt <- as.data.frame(emmeans::joint_tests(model))

  # Normalize column names
  names(jt) <- tolower(names(jt))
  names(jt) <- gsub("[[:space:]]+", "_", names(jt))
  names(jt) <- gsub("\\.", "_", names(jt))

  # Find the term column robustly
  term_col <- intersect(c("model_term", "term", "name"), names(jt))
  if (length(term_col) == 0) {
    stop("Could not find term column in joint_tests() output.")
  }
  term_col <- term_col[1]

  # Standardize output
  jt %>%
    transmute(
      Model = model_name,
      Response = response_label,
      Term = .data[[term_col]],
      `Num df` = df1,
      `Den df` = df2,
      `F ratio` = f_ratio,
      `P value raw` = p_value
    )
}

#--------------------------------------------------------
# Extract from models (edit here if you swap models)
#--------------------------------------------------------
joint_table <- bind_rows(
  extract_joint_tests(m_rdx_2,   "RDX",   "m_rdx_2"),
  extract_joint_tests(m_fdx_2,   "FDX",   "m_fdx_2"),
  extract_joint_tests(m_yield_1, "Yield (kg ha^-1)", "m_yield_1"),
  extract_joint_tests(m_yield_2, "Yield (kg ha^-1)", "m_yield_2")
)

#--------------------------------------------------------
# Clean up term labels for publication
#--------------------------------------------------------
joint_table_fmt <- joint_table %>%
  mutate(
    Term = dplyr::case_when(
      Term == "(confounded)" ~ "(confounded)",
      Term == "z_log10_fv"   ~ "F. virguliforme DNA abundance",
      Term == "z_log10_scn"  ~ "SCN density",
      Term == "z_soil_ph"    ~ "Soil pH",
      Term == "log1p_fdx"    ~ "Foliar disease index (FDX)",
      Term == "log1p_rdx"    ~ "Root disease index (RDX)",
      Term == "product"      ~ "Seed treatment",
      Term == "sds_tol"      ~ "SDS tolerance",
      TRUE ~ Term
    ),
    `Num df` = round(`Num df`, 0),
    `Den df` = round(`Den df`, 1),
    `F ratio` = round(`F ratio`, 2),
    `P value` = dplyr::case_when(
      is.na(`P value raw`) ~ "NA",
      `P value raw` < 0.0001 ~ "<0.0001",
      TRUE ~ sprintf("%.4f", `P value raw`)
    ),
    Sig. = dplyr::case_when(
      is.na(`P value raw`) ~ "",
      `P value raw` < 0.001 ~ "***",
      `P value raw` < 0.01  ~ "**",
      `P value raw` < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(
    Model,
    Response,
    Term,
    `Num df`,
    `Den df`,
    `F ratio`,
    `P value`,
    Sig.
  )

#--------------------------------------------------------
# Optional: ordering (kept simple since you're validating)
#--------------------------------------------------------
joint_table_fmt <- joint_table_fmt %>%
  mutate(
    Response = factor(
      Response,
      levels = c("FDX", "RDX", "Yield (kg ha^-1)")
    )
  ) %>%
  arrange(Model, Response) %>%
  mutate(Response = as.character(Response))

#--------------------------------------------------------
# View + export
#--------------------------------------------------------
joint_table_fmt

readr::write_csv(
  joint_table_fmt,
  "supplemental_table_final_model_joint_tests_with_model_names.csv"
)

```

### EMM table for supps ###
```{r message=FALSE, warning=FALSE}
#========================================================
# Supplemental EMM + CLD + pairwise P-value tables
#========================================================

#--------------------------------------------------------
# helper: safe response-scale back-transform for log1p models
# set transform = "log1p" when needed
#--------------------------------------------------------
backtransform_emm <- function(df, transform = c("none", "log1p")) {
  transform <- match.arg(transform)

  if (transform == "log1p") {
    df %>%
      mutate(
        emmean_bt   = exp(emmean) - 1,
        lower.CL_bt = exp(lower.CL) - 1,
        upper.CL_bt = exp(upper.CL) - 1
      )
  } else {
    df %>%
      mutate(
        emmean_bt   = emmean,
        lower.CL_bt = lower.CL,
        upper.CL_bt = upper.CL
      )
  }
}

#--------------------------------------------------------
# helper: build joined EMM summary + CLD + pairwise table
# specs examples:
#   term = "product"
#   by   = NULL
#   adjust = "sidak"
#--------------------------------------------------------
make_emm_table <- function(model,
                           model_name,
                           response_label,
                           term,
                           by = NULL,
                           adjust = "sidak",
                           transform = c("none", "log1p"),
                           digits = 2) {

  transform <- match.arg(transform)

  # EMM object
  emm <- if (is.null(by)) {
    emmeans::emmeans(model, specs = as.formula(paste("~", term)))
  } else {
    emmeans::emmeans(model, specs = as.formula(paste("~", term, "|", by)))
  }

  # summary table
  emm_df <- as.data.frame(summary(emm, infer = c(TRUE, TRUE))) %>%
    backtransform_emm(transform = transform)

  # CLD letters
  cld_df <- as.data.frame(
    multcomp::cld(emm, adjust = adjust, Letters = letters)
  ) %>%
    mutate(.group = str_trim(.group))

  # pairwise comparisons with exact adjusted p-values
  pair_df <- as.data.frame(
    summary(pairs(emm, adjust = adjust), infer = c(TRUE, TRUE))
  )

  # join CLD to EMM table
  join_cols <- intersect(names(emm_df), names(cld_df))
  emm_out <- emm_df %>%
    left_join(
      cld_df %>% select(all_of(join_cols), .group),
      by = join_cols
    ) %>%
    mutate(
      Model = model_name,
      Response = response_label,
      Term = term,
      By = ifelse(is.null(by), NA_character_, by),
      emmean_bt   = round(emmean_bt, digits),
      lower.CL_bt = round(lower.CL_bt, digits),
      upper.CL_bt = round(upper.CL_bt, digits)
    )

  # format p-values in pairwise table
  pair_out <- pair_df %>%
    mutate(
      Model = model_name,
      Response = response_label,
      Term = term,
      By = ifelse(is.null(by), NA_character_, by),
      p.value_fmt = case_when(
        is.na(p.value) ~ "NA",
        p.value < 0.0001 ~ "<0.0001",
        TRUE ~ sprintf("%.4f", p.value)
      )
    )

  list(
    emm = emm_out,
    pairs = pair_out
  )
}

#========================================================
# EXAMPLES: build targeted tables
# Edit these model calls as needed
#========================================================

#-----------------------------
# RDX
#-----------------------------
rdx_product <- make_emm_table(
  model = m_rdx_2,
  model_name = "m_rdx_2",
  response_label = "RDX",
  term = "product",
  adjust = "sidak",
  transform = "log1p"
)

rdx_tol <- make_emm_table(
  model = m_rdx_2,
  model_name = "m_rdx_2",
  response_label = "RDX",
  term = "sds_tol",
  adjust = "sidak",
  transform = "log1p"
)

# only use this if the interaction is in the model and you want sliced results
rdx_product_by_tol <- make_emm_table(
   model = m_rdx_2,
   model_name = "m_rdx_2",
   response_label = "RDX",
   term = "product",
   by = "sds_tol",
   adjust = "sidak",
   transform = "log1p"
 )

#-----------------------------
# FDX
#-----------------------------
fdx_product <- make_emm_table(
  model = m_fdx_2,
  model_name = "m_fdx_2",
  response_label = "FDX",
  term = "product",
  adjust = "sidak",
  transform = "log1p"
)

fdx_tol <- make_emm_table(
  model = m_fdx_2,
  model_name = "m_fdx_2",
  response_label = "FDX",
  term = "sds_tol",
  adjust = "sidak",
  transform = "log1p"
)

# only use this if the interaction is in the model and you want sliced results
fdx_product_by_tol <- make_emm_table(
   model = m_fdx_2,
   model_name = "m_fdx_2",
   response_label = "FDX",
   term = "product",
   by = "sds_tol",
   adjust = "sidak",
   transform = "log1p"
 )

#-----------------------------
# Yield
#-----------------------------
yield1_product <- make_emm_table(
  model = m_yield_1,
  model_name = "m_yield_1",
  response_label = "Yield (with disease indices)",
  term = "product",
  adjust = "sidak",
  transform = "none"
)

yield1_tol <- make_emm_table(
  model = m_yield_1,
  model_name = "m_yield_1",
  response_label = "Yield (with disease indices)",
  term = "sds_tol",
  adjust = "sidak",
  transform = "none"
)

# use this if the interaction is in the model and you want the actionable slices
yield1_product_by_tol <- make_emm_table(
  model = m_yield_1,
  model_name = "m_yield_1",
  response_label = "Yield (with disease indices)",
  term = "product",
  by = "sds_tol",
  adjust = "sidak",
  transform = "none"
)

yield2_product <- make_emm_table(
  model = m_yield_2,
  model_name = "m_yield_2",
  response_label = "Yield (without disease indices)",
  term = "product",
  adjust = "sidak",
  transform = "none"
)

yield2_tol <- make_emm_table(
  model = m_yield_2,
  model_name = "m_yield_2",
  response_label = "Yield (without disease indices)",
  term = "sds_tol",
  adjust = "sidak",
  transform = "none"
)

# use this if the interaction is in the model and you want the actionable slices
yield2_product_by_tol <- make_emm_table(
  model = m_yield_2,
  model_name = "m_yield_2",
  response_label = "Yield (without disease indices)",
  term = "product",
  by = "sds_tol",
  adjust = "sidak",
  transform = "none"
)

#========================================================
# bind outputs
#========================================================

emm_summary_table <- bind_rows(
  rdx_product$emm,
  rdx_tol$emm,
  rdx_product_by_tol$emm,
  fdx_product$emm,
  fdx_tol$emm,
  fdx_product_by_tol$emm,
  yield1_product$emm,
  yield1_tol$emm,
  yield1_product_by_tol$emm,
  yield2_product$emm,
  yield2_tol$emm,
  yield2_product_by_tol$emm
)

pairwise_p_table <- bind_rows(
  rdx_product$pairs,
  rdx_tol$pairs,
  rdx_product_by_tol$pairs,
  fdx_product$pairs,
  fdx_tol$pairs,
  fdx_product_by_tol$pairs,
  yield1_product$pairs,
  yield1_tol$pairs,
  yield1_product_by_tol$pairs,
  yield2_product$pairs,
  yield2_tol$pairs,
  yield2_product_by_tol$pairs
)

#========================================================
# optional cleanup for Word export
#========================================================

emm_summary_table_fmt <- emm_summary_table %>%
  mutate(
    across(any_of(c("product", "sds_tol")), as.character),
    CLD = .group
  ) %>%
  select(
    Model, Response, Term, By,
    any_of(c("product", "sds_tol")),
    emmean_bt, lower.CL_bt, upper.CL_bt, CLD
  )

pairwise_p_table_fmt <- pairwise_p_table %>%
  mutate(
    contrast = as.character(contrast)
  ) %>%
  select(
    Model, Response, Term, By, contrast, estimate, SE, df, t.ratio, p.value_fmt
  )

#========================================================
# export
#========================================================

readr::write_csv(
  emm_summary_table_fmt,
  "supplemental_emm_summary_with_cld.csv"
)

readr::write_csv(
  pairwise_p_table_fmt,
  "supplemental_emm_pairwise_pvalues.csv"
)
```


### surface response plots by product ###
```{r warning=FALSE}
library(plotly)


product_levels <- c("Base", "ILEVO", "Saltro")
# If needed, use:
# product_levels <- c("CB", "ILEVO", "Saltro")

surf_df <- df %>%
  mutate(
    fv_log    = log10(ratio_fvfg_dnang + 1),
    scn_log   = log10(scn_count_rep + 1),
    rdx_log1p = log1p(rdx),
    fdx_log1p = log1p(fdx)
  ) %>%
  filter(product %in% product_levels)

range_x <- range(surf_df$fv_log, na.rm = TRUE)
range_y <- range(surf_df$scn_log, na.rm = TRUE)

pred_grid <- expand.grid(
  fv_log  = seq(range_x[1], range_x[2], length.out = 100),
  scn_log = seq(range_y[1], range_y[2], length.out = 100)
)

fit_product_surface <- function(dat, response) {
  if (response == "RDX") {
    dat_use <- dat %>% filter(!is.na(rdx_log1p), !is.na(fv_log), !is.na(scn_log))
    mod <- gam(rdx_log1p ~ s(fv_log, scn_log, k = 20), data = dat_use, method = "REML")
    pred <- pred_grid %>%
      mutate(
        fit = predict(mod, newdata = pred_grid, type = "response"),
        response = exp(fit) - 1
      )
  } else if (response == "FDX") {
    dat_use <- dat %>% filter(!is.na(fdx_log1p), !is.na(fv_log), !is.na(scn_log))
    mod <- gam(fdx_log1p ~ s(fv_log, scn_log, k = 20), data = dat_use, method = "REML")
    pred <- pred_grid %>%
      mutate(
        fit = predict(mod, newdata = pred_grid, type = "response"),
        response = exp(fit) - 1
      )
  } else if (response == "Yield") {
    dat_use <- dat %>% filter(!is.na(yield_kgha), !is.na(fv_log), !is.na(scn_log))
    mod <- gam(yield_kgha ~ s(fv_log, scn_log, k = 20), data = dat_use, method = "REML")
    pred <- pred_grid %>%
      mutate(
        response = predict(mod, newdata = pred_grid, type = "response")
      )
  } else {
    stop("response must be one of: RDX, FDX, Yield")
  }

  list(model = mod, pred = pred, obs = dat_use)
}

make_surface_matrix <- function(pred_dat) {
  x_vals <- sort(unique(pred_dat$fv_log))
  y_vals <- sort(unique(pred_dat$scn_log))

  z_mat <- pred_dat %>%
    arrange(scn_log, fv_log) %>%
    select(scn_log, fv_log, response) %>%
    pivot_wider(names_from = fv_log, values_from = response) %>%
    select(-scn_log) %>%
    as.matrix()

  list(x = x_vals, y = y_vals, z = z_mat)
}

fits <- list()

for (p in product_levels) {
  dat_p <- surf_df %>% filter(product == p)

  fits[[paste0("RDX_", p)]]   <- fit_product_surface(dat_p, "RDX")
  fits[[paste0("FDX_", p)]]   <- fit_product_surface(dat_p, "FDX")
  fits[[paste0("Yield_", p)]] <- fit_product_surface(dat_p, "Yield")
}

surfaces <- list()

for (nm in names(fits)) {
  surfaces[[nm]] <- make_surface_matrix(fits[[nm]]$pred)
}

zlim_rdx <- range(unlist(lapply(product_levels, function(p) surfaces[[paste0("RDX_", p)]]$z)), na.rm = TRUE)
zlim_fdx <- range(unlist(lapply(product_levels, function(p) surfaces[[paste0("FDX_", p)]]$z)), na.rm = TRUE)
zlim_yld <- range(unlist(lapply(product_levels, function(p) surfaces[[paste0("Yield_", p)]]$z)), na.rm = TRUE)

zlim_yld[1] <- max(0, zlim_yld[1] - 300)

```

#yeh
```{r}

camera_eye <- list(x = 1.25, y = 1, z = 1.75) #sets camera eye globally and assigns it to a list

scene_common <- function(ztitle, range_x, range_y, zlim) {
  list(
    xaxis = list(
      title = list(text = "log10(F. virguliforme (fg•ng))", font = list(size = 12)),
      tickfont = list(size = 12),
      range = range_x,
      showbackground = TRUE,
      backgroundcolor = "white",
      gridcolor = "gray20",
      zerolinecolor = "black"
    ),
    yaxis = list(
      title = list(text = "log10(SCN eggs + J2s)", font = list(size = 12)),
      tickfont = list(size = 12),
      range = range_y,
      showbackground = TRUE,
      backgroundcolor = "white",
      gridcolor = "gray20",
      zerolinecolor = "black"
    ),
    zaxis = list(
      title = list(text = ztitle, font = list(size = 12)),
      tickfont = list(size = 12),
      range = zlim,
      showbackground = TRUE,
      backgroundcolor = "white",
      gridcolor = "gray20",
      zerolinecolor = "black"
    ),
    camera = camera_eye,
    aspectratio = list(x = 1, y = 1, z = 0.75)
  )
}

make_surface_plot <- function(surfmat, obs_dat, ztitle, zlim, range_x, range_y,
                              colorscale = "Viridis",
                              show_scale = FALSE,
                              opacity = 0.30,
                              mesh_color = "gray30",
                              mesh_width = 0.25,
                              point_color = "black",
                              point_size = 1.2,
                              point_frac = 0.35,
                              show_points = TRUE) {

  z_floor <- zlim[1] - 0.05 * diff(zlim)

  p <- plot_ly() %>%
    add_surface(
      x = surfmat$x,
      y = surfmat$y,
      z = surfmat$z,
      colorscale = colorscale,
      showscale = show_scale,
      showpoints = show_points,
      opacity = opacity,
      contours = list(
        x = list(show = TRUE, color = mesh_color, width = mesh_width, start = min(range_x),
        end = max(range_x),
        size = diff(range_x) / 20),
        y = list(show = TRUE, color = mesh_color, width = mesh_width, start = min(range_y),
        end = max(range_y),
        size = diff(range_y) / 20),
        z = list(show = FALSE)
      ),
      hovertemplate = paste0(
        "log10(Fv): %{x:.2f}<br>",
        "log10(SCN): %{y:.2f}<br>",
        ztitle, ": %{z:.2f}<extra></extra>"
      )
    )

  if (show_points) {
    obs_use <- obs_dat %>% sample_frac(min(1, point_frac))
    p <- p %>%
      add_markers(
        data = obs_use,
        x = ~fv_log,
        y = ~scn_log,
        z = rep(z_floor, nrow(obs_use)),
        marker = list(
          size = point_size,
          color = point_color,
          opacity = 0.40,
          symbol = "circle"
        ),
        inherit = TRUE,
        hovertemplate = paste0(
          "Observed<br>",
          "log10(Fv): %{x:.2f}<br>",
          "log10(SCN): %{y:.2f}<extra></extra>"
        ),
        showlegend = FALSE
      )
  }

  p %>%
    layout(
      font = list(size = 12),
      margin = list(l = 0, r = 0, b = 0, t = 0),
      scene = scene_common(ztitle, range_x, range_y, zlim)
      )
}
```

#uh huh
```{r}
fig_rdx_base <- make_surface_plot(
  surfmat = surfaces[["RDX_Base"]],
  obs_dat = fits[["RDX_Base"]]$obs,
  ztitle = "Root disease index (RDX)",
  zlim = zlim_rdx,
  range_x = range_x,
  range_y = range_y,
)

fig_rdx_ilevo <- make_surface_plot(
  surfmat = surfaces[["RDX_ILEVO"]],
  obs_dat = fits[["RDX_ILEVO"]]$obs,
  ztitle = "Root disease index (RDX)",
  zlim = zlim_rdx,
  range_x = range_x,
  range_y = range_y
)

fig_rdx_saltro <- make_surface_plot(
  surfmat = surfaces[["RDX_Saltro"]],
  obs_dat = fits[["RDX_Saltro"]]$obs,
  ztitle = "Root disease index (RDX)",
  zlim = zlim_rdx,
  range_x = range_x,
  range_y = range_y
)

fig_fdx_base <- make_surface_plot(
  surfmat = surfaces[["FDX_Base"]],
  obs_dat = fits[["FDX_Base"]]$obs,
  ztitle = "Foliar disease index (FDX)",
  zlim = zlim_fdx,
  range_x = range_x,
  range_y = range_y
)

fig_fdx_ilevo <- make_surface_plot(
  surfmat = surfaces[["FDX_ILEVO"]],
  obs_dat = fits[["FDX_ILEVO"]]$obs,
  ztitle = "Foliar disease index (FDX)",
  zlim = zlim_fdx,
  range_x = range_x,
  range_y = range_y
)

fig_fdx_saltro <- make_surface_plot(
  surfmat = surfaces[["FDX_Saltro"]],
  obs_dat = fits[["FDX_Saltro"]]$obs,
  ztitle = "Foliar disease index (FDX)",
  zlim = zlim_fdx,
  range_x = range_x,
  range_y = range_y
)

fig_yld_base <- make_surface_plot(
  surfmat = surfaces[["Yield_Base"]],
  obs_dat = fits[["Yield_Base"]]$obs,
  ztitle = "Soybean yield (kg•hectare)",
  zlim = zlim_yld,
  range_x = range_x,
  range_y = range_y
)

fig_yld_ilevo <- make_surface_plot(
  surfmat = surfaces[["Yield_ILEVO"]],
  obs_dat = fits[["Yield_ILEVO"]]$obs,
  ztitle = "Soybean yield (kg•hectare)",
  zlim = zlim_yld,
  range_x = range_x,
  range_y = range_y
)

fig_yld_saltro <- make_surface_plot(
  surfmat = surfaces[["Yield_Saltro"]],
  obs_dat = fits[["Yield_Saltro"]]$obs,
  ztitle = "Soybean yield (kg•hectare)",
  zlim = zlim_yld,
  range_x = range_x,
  range_y = range_y
)
```

#yupper
```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#saveWidget(fig_rdx_base,   "panel_rdx_base.html",   selfcontained = TRUE)
#saveWidget(fig_rdx_ilevo,  "panel_rdx_ilevo.html",  selfcontained = TRUE)
#saveWidget(fig_rdx_saltro, "panel_rdx_saltro.html", selfcontained = TRUE)

#saveWidget(fig_fdx_base,   "panel_fdx_base.html",   selfcontained = TRUE)
#saveWidget(fig_fdx_ilevo,  "panel_fdx_ilevo.html",  selfcontained = TRUE)
#saveWidget(fig_fdx_saltro, "panel_fdx_saltro.html", selfcontained = TRUE)

#saveWidget(fig_yld_base,   "panel_yld_base.html",   selfcontained = TRUE)
#saveWidget(fig_yld_ilevo,  "panel_yld_ilevo.html",  selfcontained = TRUE)
#saveWidget(fig_yld_saltro, "panel_yld_saltro.html", selfcontained = TRUE)

```

#ok
```{r message=FALSE, warning=FALSE}
#install.packages("webshot2")
#webshot2::install_phantomjs()

library(webshot2)

save_plotly_png <- function(fig, file, width = 600, height = 500, zoom = 3) {
  
  tmp_html <- tempfile(fileext = ".html")
  
  htmlwidgets::saveWidget(fig, tmp_html, selfcontained = TRUE)
  
  webshot2::webshot(
    url = tmp_html,
    file = file,
    vwidth = width,
    vheight = height,
    zoom = zoom
  )
}

save_plotly_png(fig_rdx_base,   "panel_rdx_base.png")
save_plotly_png(fig_rdx_ilevo,  "panel_rdx_ilevo.png")
save_plotly_png(fig_rdx_saltro, "panel_rdx_saltro.png")

save_plotly_png(fig_fdx_base,   "panel_fdx_base.png")
save_plotly_png(fig_fdx_ilevo,  "panel_fdx_ilevo.png")
save_plotly_png(fig_fdx_saltro, "panel_fdx_saltro.png")

save_plotly_png(fig_yld_base,   "panel_yld_base.png")
save_plotly_png(fig_yld_ilevo,  "panel_yld_ilevo.png")
save_plotly_png(fig_yld_saltro, "panel_yld_saltro.png")
```



#yah
```{r}
library(cowplot)
library(magick)

label_panel <- function(path, lab,
                        x = -0.1, y = 0.8,
                        size = 12) {
  ggdraw() +
    draw_image(path) +
    draw_label(
      lab,
      x = x, y = y,
      hjust = 0, vjust = 0,
      fontface = "bold",
      size = size
    )
}

p1 <- label_panel("panel_rdx_base.png",   "A1")
p2 <- label_panel("panel_rdx_ilevo.png",  "A2")
p3 <- label_panel("panel_rdx_saltro.png", "A3")

p4 <- label_panel("panel_fdx_base.png",   "B1")
p5 <- label_panel("panel_fdx_ilevo.png",  "B2")
p6 <- label_panel("panel_fdx_saltro.png", "B3")

p7 <- label_panel("panel_yld_base.png",   "C1")
p8 <- label_panel("panel_yld_ilevo.png",  "C2")
p9 <- label_panel("panel_yld_saltro.png", "C3")


row1 <- plot_grid(p1, p2, p3, ncol = 3, align = "hv", axis = "tblr")
row2 <- plot_grid(p4, p5, p6, ncol = 3, align = "hv", axis = "tblr")
row3 <- plot_grid(p7, p8, p9, ncol = 3, align = "hv", axis = "tblr")

col_titles <- plot_grid(
  ggdraw() + draw_label("Commercial Base (CB)", fontface = "bold", size = 12,y=-0.8),
  ggdraw() + draw_label("CB + fluopyram", fontface = "bold", size = 12,y=-0.8),
  ggdraw() + draw_label("CB + pydiflumetofen", fontface = "bold", size = 12,y=-0.8),
  ncol = 3
)

row1 <- plot_grid(p1, p2, p3, ncol = 3, align = "hv")
row2 <- plot_grid(p4, p5, p6, ncol = 3, align = "hv")
row3 <- plot_grid(p7, p8, p9, ncol = 3, align = "hv")

body <- plot_grid(
  ggdraw() + draw_label("Root disease", angle = 90, fontface = "bold", size = 12),
  row1,
  ggdraw() + draw_label("Foliar disease", angle = 90, fontface = "bold", size = 12),
  row2,
  ggdraw() + draw_label("Yield", angle = 90, fontface = "bold", size = 12),
  row3,
  ncol = 2,
  rel_widths = c(0.06, 1),
  rel_heights = c(1, 0.7, 1, 0.9, 1, 0.9)  # equal, no padding bias
)

final_fig <- plot_grid(
  NULL, col_titles,
  NULL, body,
  ncol = 2,
  rel_widths = c(0.06, 1),
  rel_heights = c(0.035, 1)
)

ggsave("figure_9panel_products.png", final_fig, width = 12, height = 12, dpi = 600)
```


#### ROI code for summaries ####
```{r include=FALSE}

#========================================================
# Decision-scenario yield and ROI extraction
# Risk class × product × SDS tolerance decision model
#========================================================

#-----------------------------
# User settings
#-----------------------------
treatment_cost_usd_ha <- 32
price_usd_kg <- 0.45

#-----------------------------
# Helpers
#-----------------------------
standardize_product <- function(x) {
  x <- as.character(x)
  x <- stringr::str_trim(x)

  dplyr::case_when(
    x %in% c("Base", "CB", "Commercial Base", "Commercial base", "Commercial base (CB)") ~ "Base",
    stringr::str_detect(x, stringr::regex("ILEVO|fluopyram", ignore_case = TRUE)) ~ "ILEVO",
    stringr::str_detect(x, stringr::regex("Saltro|pydiflumetofen", ignore_case = TRUE)) ~ "Saltro",
    TRUE ~ x
  )
}

align_factor_to_model <- function(x, model, varname) {
  mf <- model.frame(model)

  if (varname %in% names(mf) && is.factor(mf[[varname]])) {
    factor(as.character(x), levels = levels(mf[[varname]]))
  } else {
    x
  }
}

get_center <- function(x) mean(x, na.rm = TRUE)
get_scale  <- function(x) sd(x, na.rm = TRUE)

#-----------------------------
# 1. Build risk classes in model dataframe
#-----------------------------
df <- df_yield_mod %>%
  mutate(
    product = factor(
      standardize_product(product),
      levels = c("Base", "ILEVO", "Saltro")
    ),
    sds_tol = factor(sds_tol),

    fv_level = case_when(
      ratio_fvfg_dnang == 0 ~ "0",
      ratio_fvfg_dnang > 0 & ratio_fvfg_dnang < 1 ~ "0-1",
      ratio_fvfg_dnang >= 1 & ratio_fvfg_dnang < 3 ~ "1-3",
      ratio_fvfg_dnang >= 3 & ratio_fvfg_dnang < 5 ~ "3-5",
      ratio_fvfg_dnang >= 5 ~ ">5",
      TRUE ~ NA_character_
    ),
    fv_level = factor(
      fv_level,
      levels = c("0", "0-1", "1-3", "3-5", ">5"),
      ordered = TRUE
    ),

    scn_level = case_when(
      scn_count_rep == 0 ~ "None",
      scn_count_rep < 2000 ~ "Low",
      scn_count_rep >= 2000 ~ "High",
      TRUE ~ NA_character_
    ),
    scn_level = factor(
      scn_level,
      levels = c("None", "Low", "High"),
      ordered = TRUE
    ),

    fdx_level = case_when(
      fdx == 0 ~ "None",
      fdx > 0 & fdx < 10 ~ "Low",
      fdx >= 10 ~ "High",
      TRUE ~ NA_character_
    ),
    fdx_level = factor(
      fdx_level,
      levels = c("None", "Low", "High"),
      ordered = TRUE
    ),

    risk_class = interaction(fv_level, scn_level, drop = TRUE),
    risk_class = droplevels(factor(risk_class))
  ) %>%
  filter(
    !is.na(product),
    !is.na(sds_tol),
    !is.na(risk_class),
    !is.na(yield_kgha),
    !is.na(z_soil_ph),
    !is.na(siteyear),
    !is.na(exp_siteyear)
  )

#-----------------------------
# 2. Fit decision model
#-----------------------------
m_yield_decision <- lmer(
  yield_kgha ~
    product * risk_class +
    product * sds_tol +
    z_soil_ph +
    (1 | siteyear) +
    (1 | exp_siteyear),
  data = df,
  REML = FALSE,
  control = lmerControl(
    optimizer = "bobyqa",
    optCtrl = list(maxfun = 200000)
  )
)

#emmeans::joint_tests(m_yield_decision)

#-----------------------------
# 3. Cell support diagnostics
#-----------------------------
cell_support <- df %>%
  count(risk_class, sds_tol, product, name = "n") %>%
  arrange(risk_class, sds_tol, product)

supported_cells <- cell_support %>%
  group_by(risk_class, sds_tol) %>%
  summarise(
    n_products = n_distinct(product),
    min_n = min(n),
    total_n = sum(n),
    .groups = "drop"
  ) %>%
  filter(
    n_products == 3,
    min_n >= 3,
    total_n >= 12
  )

unsupported_pairs <- cell_support %>%
  group_by(risk_class, sds_tol) %>%
  summarise(
    n_products = n_distinct(product),
    min_n = min(n),
    total_n = sum(n),
    .groups = "drop"
  ) %>%
  filter(
    n_products < 3 | min_n < 3 | total_n < 12
  )

valid_keys <- supported_cells %>%
  select(risk_class, sds_tol)

#-----------------------------
# 4. EMMs and pairwise contrasts
#-----------------------------
emm_decision <- emmeans(
  m_yield_decision,
  ~ product | risk_class * sds_tol
)

pairs_decision <- pairs(
  emm_decision,
  adjust = "tukey"
)

pairs_decision_df <- as.data.frame(pairs_decision) %>%
  filter(!is.na(p.value)) %>%
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

pairs_clean <- pairs_decision_df %>%
  semi_join(valid_keys, by = c("risk_class", "sds_tol"))

pairs_decision_sig <- pairs_clean %>%
  filter(p.value < 0.05)

#-----------------------------
# 5. Scaling constants for decision scenarios
#-----------------------------
df_scale <- df %>%
  mutate(
    fv_log = log10(ratio_fvfg_dnang + 1),
    scn_log = log10(scn_count_rep + 1)
  )

fv_center  <- get_center(df_scale$fv_log)
fv_scale   <- get_scale(df_scale$fv_log)
scn_center <- get_center(df_scale$scn_log)
scn_scale  <- get_scale(df_scale$scn_log)
ph_center  <- get_center(df_scale$soil_ph)
ph_scale   <- get_scale(df_scale$soil_ph)

model_risk_levels <- levels(model.frame(m_yield_decision)$risk_class)


model_risk_levels

#-----------------------------
# 6. Scenario addresses
#-----------------------------
decision_scenarios <- tibble::tribble(
  ~scenario,                         ~fv_fg_ng, ~scn_eggs_j2_100cc, ~fdx, ~rdx, ~soil_ph,
  "No detectable risk",               0.0,       0,                  0,    0,    ph_center,
  "Fv 0-1 / No SCN / No FDX",          0.5,       0,                  0,    0,    ph_center,
  "Fv 1-3 / Low SCN / Low FDX",        2.0,       1000,               5,    5,    ph_center,
  "Fv 3-5 / High SCN / Low FDX",       4.0,       3000,               5,    10,   ph_center,
  "Fv >5 / Low SCN / No FDX",          6.0,       1000,               0,    5,    ph_center,
  "Fv >5 / High SCN / High FDX",       6.0,       3000,               10,   15,   ph_center
) %>%
  mutate(
    fv_log = log10(fv_fg_ng + 1),
    scn_log = log10(scn_eggs_j2_100cc + 1),

    z_log10_fv  = (fv_log - fv_center) / fv_scale,
    z_log10_scn = (scn_log - scn_center) / scn_scale,
    z_soil_ph   = (soil_ph - ph_center) / ph_scale,

    log1p_fdx = log1p(fdx),
    log1p_rdx = log1p(rdx),

    fv_level = case_when(
      fv_fg_ng == 0 ~ "0",
      fv_fg_ng > 0 & fv_fg_ng < 1 ~ "0-1",
      fv_fg_ng >= 1 & fv_fg_ng < 3 ~ "1-3",
      fv_fg_ng >= 3 & fv_fg_ng < 5 ~ "3-5",
      fv_fg_ng >= 5 ~ ">5",
      TRUE ~ NA_character_
    ),
    fv_level = factor(
      fv_level,
      levels = c("0", "0-1", "1-3", "3-5", ">5"),
      ordered = TRUE
    ),

    scn_level = case_when(
      scn_eggs_j2_100cc == 0 ~ "None",
      scn_eggs_j2_100cc < 2000 ~ "Low",
      scn_eggs_j2_100cc >= 2000 ~ "High",
      TRUE ~ NA_character_
    ),
    scn_level = factor(
      scn_level,
      levels = c("None", "Low", "High"),
      ordered = TRUE
    ),

    fdx_level = case_when(
      fdx == 0 ~ "None",
      fdx > 0 & fdx < 10 ~ "Low",
      fdx >= 10 ~ "High",
      TRUE ~ NA_character_
    ),
    fdx_level = factor(
      fdx_level,
      levels = c("None", "Low", "High"),
      ordered = TRUE
    ),

    risk_class = interaction(fv_level, scn_level, drop = TRUE)
  )

# Align scenario risk classes to model-supported levels
model_risk_levels <- levels(model.frame(m_yield_decision)$risk_class)

decision_scenarios_check <- decision_scenarios %>%
  mutate(
    risk_class_chr = as.character(risk_class),
    in_model = risk_class_chr %in% model_risk_levels
  ) %>%
  select(scenario, fv_level, scn_level, risk_class_chr, in_model)

decision_scenarios_check

model_risk_levels

#-----------------------------
# 7. Product × tolerance prediction grid
#-----------------------------
product_levels_decision <- levels(model.frame(m_yield_decision)$product)
sds_tol_levels_decision <- levels(model.frame(m_yield_decision)$sds_tol)

decision_grid <- tidyr::crossing(
  decision_scenarios,
  product = product_levels_decision,
  sds_tol = sds_tol_levels_decision
)

#-----------------------------
# 8. Predict from decision model
#-----------------------------
predict_decision_model <- function(model, model_name, newdat) {

  nd <- newdat %>%
    mutate(
      product = align_factor_to_model(product, model, "product"),
      sds_tol = align_factor_to_model(sds_tol, model, "sds_tol"),
      risk_class = align_factor_to_model(risk_class, model, "risk_class")
    )

  nd$pred_yield_kgha <- as.numeric(
    predict(model, newdata = nd, re.form = NA, allow.new.levels = TRUE)
  )

  nd %>% mutate(model = model_name)
}

yield_models <- list(
  m_yield_decision = m_yield_decision
)

decision_predictions <- purrr::imap_dfr(
  yield_models,
  ~ predict_decision_model(.x, .y, decision_grid)
)

#-----------------------------
# 9. ROI calculation
#-----------------------------
decision_roi <- decision_predictions %>%
  group_by(model, risk_class, scenario, sds_tol) %>%
  mutate(
    base_yield_kgha = pred_yield_kgha[as.character(product) == "Base"][1],
    delta_yield_kgha = pred_yield_kgha - base_yield_kgha,
    incremental_cost_usd_ha = if_else(as.character(product) == "Base", 0, treatment_cost_usd_ha),
    gross_return_usd_ha = delta_yield_kgha * price_usd_kg,
    net_return_usd_ha = gross_return_usd_ha - incremental_cost_usd_ha,
    break_even_kgha = treatment_cost_usd_ha / price_usd_kg,
    roi_status = case_when(
      as.character(product) == "Base" ~ "Reference",
      net_return_usd_ha > 0 ~ "Positive ROI",
      net_return_usd_ha <= 0 ~ "Negative/neutral ROI",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup()

decision_roi_summary_fmt <- decision_roi %>%
  select(
    model, scenario, risk_class,
    fv_level, scn_level, fdx_level,
    fv_fg_ng, scn_eggs_j2_100cc, fdx, rdx, soil_ph,
    sds_tol, product,
    pred_yield_kgha,
    delta_yield_kgha,
    gross_return_usd_ha,
    incremental_cost_usd_ha,
    net_return_usd_ha,
    break_even_kgha,
    roi_status
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

roi_clean <- decision_roi_summary_fmt %>%
  semi_join(valid_keys, by = c("risk_class", "sds_tol"))

#-----------------------------
# 10. Export
#-----------------------------
output_dir <- "decision_scenario_outputs"

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

utils::write.csv(
  decision_roi_summary_fmt,
  file.path(output_dir, "decision_scenario_yield_roi_summary_with_cb.csv"),
  row.names = FALSE
)

utils::write.csv(
  roi_clean,
  file.path(output_dir, "decision_scenario_yield_roi_supported_only.csv"),
  row.names = FALSE
)

utils::write.csv(
  pairs_clean,
  file.path(output_dir, "decision_scenario_pairwise_supported_only.csv"),
  row.names = FALSE
)

utils::write.csv(
  pairs_decision_sig,
  file.path(output_dir, "decision_scenario_pairwise_significant.csv"),
  row.names = FALSE
)

utils::write.csv(
  cell_support,
  file.path(output_dir, "decision_scenario_cell_support.csv"),
  row.names = FALSE
)

utils::write.csv(
  unsupported_pairs,
  file.path(output_dir, "decision_scenario_unsupported_pairs.csv"),
  row.names = FALSE
)

#-----------------------------
# 11. Print key outputs
#-----------------------------
supported_cells
unsupported_pairs
roi_clean
pairs_clean
pairs_decision_sig


```

### ROI plot and tables ###
```{r}

#========================================================
# Main figure + supplemental table for decision ROI results
# Requires existing objects:
#   decision_roi_summary_fmt
#   pairs_clean
#   pairs_decision_sig
#========================================================

#-----------------------------
# 1. Clean and order ROI table robustly
#-----------------------------

emm_df <- as.data.frame(emm_decision)

decision_roi_summary_fmt <- emm_df %>%
  group_by(risk_class, sds_tol) %>%
  mutate(
    base_yield_kgha = emmean[as.character(product) == "Base"][1],
    delta_yield_kgha = emmean - base_yield_kgha,
    incremental_cost_usd_ha = if_else(as.character(product) == "Base", 0, treatment_cost_usd_ha),
    gross_return_usd_ha = delta_yield_kgha * price_usd_kg,
    net_return_usd_ha = gross_return_usd_ha - incremental_cost_usd_ha,
    break_even_kgha = treatment_cost_usd_ha / price_usd_kg,
    roi_status = case_when(
      as.character(product) == "Base" ~ "Reference",
      net_return_usd_ha > 0 ~ "Positive ROI",
      net_return_usd_ha <= 0 ~ "Negative/neutral ROI",
      TRUE ~ NA_character_
    )
  ) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

decision_roi_summary_fmt

# Inspect actual levels first
unique(decision_roi_summary_fmt$risk_class)
unique(decision_roi_summary_fmt$product)
unique(decision_roi_summary_fmt$sds_tol)

risk_order <- decision_roi_summary_fmt %>%
  dplyr::mutate(risk_class = as.character(risk_class)) %>%
  dplyr::distinct(risk_class) %>%
  dplyr::pull(risk_class)

decision_roi_summary_fmt %>%
  dplyr::summarise(
    n = dplyr::n(),
    products = paste(unique(as.character(product)), collapse = " | "),
    sds_tols = paste(unique(as.character(sds_tol)), collapse = " | "),
    risk_classes = paste(unique(as.character(risk_class)), collapse = " | ")
  )

roi_plot_df <- decision_roi_summary_fmt %>%
  dplyr::mutate(
    risk_class = as.character(risk_class),
    product_raw = as.character(product),
    sds_tol_raw = as.character(sds_tol),

    product_clean = dplyr::case_when(
      stringr::str_detect(product_raw, stringr::regex("base|CB$", ignore_case = TRUE)) ~ "Base",
      stringr::str_detect(product_raw, stringr::regex("ILEVO|fluopyram", ignore_case = TRUE)) ~ "CB + fluopyram",
      stringr::str_detect(product_raw, stringr::regex("Saltro|pydiflumetofen", ignore_case = TRUE)) ~ "CB + pydiflumetofen",
      TRUE ~ product_raw
    ),

    sds_tol_clean = dplyr::case_when(
      sds_tol_raw %in% c("r", "R", "resistant", "Resistant") ~ "Moderately resistant cultivar",
      sds_tol_raw %in% c("s", "S", "susceptible", "Susceptible") ~ "Susceptible cultivar",
      TRUE ~ sds_tol_raw
    )
  ) %>%
  dplyr::filter(product_clean != "Base") %>%
  dplyr::mutate(
    risk_class = factor(risk_class, levels = unique(risk_class)),
    product = factor(
      product_clean,
      levels = c("CB + fluopyram", "CB + pydiflumetofen")
    ),
    sds_tol = factor(
      sds_tol_clean,
      levels = c("Moderately resistant cultivar", "Susceptible cultivar")
    )
  ) %>%
  dplyr::filter(!is.na(product), !is.na(sds_tol), !is.na(risk_class))

# Check before plotting
nrow(roi_plot_df)
dplyr::count(roi_plot_df, risk_class, sds_tol, product)

if (nrow(roi_plot_df) == 0) {
  stop("roi_plot_df is empty. Check product, sds_tol, and risk_class labels.")
}

#-----------------------------
# 2. Add significance flags
#-----------------------------
sig_flags <- pairs_decision_sig %>%
  dplyr::mutate(
    risk_class = as.character(risk_class),
    sds_tol = factor(
      as.character(sds_tol),
      levels = c("r", "s"),
      labels = c("Moderately resistant cultivar", "Susceptible cultivar")
    ),
    contrast = as.character(contrast),
    sig_label = dplyr::case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  dplyr::filter(stringr::str_detect(contrast, "Saltro")) %>%
  dplyr::group_by(risk_class, sds_tol) %>%
  dplyr::summarise(
    sig_label = paste(unique(sig_label), collapse = ""),
    .groups = "drop"
  )

roi_plot_df <- roi_plot_df %>%
  dplyr::left_join(sig_flags, by = c("risk_class", "sds_tol")) %>%
  dplyr::mutate(
    highlight = dplyr::if_else(
      as.character(product) == "CB + pydiflumetofen" &
        !is.na(sig_label) &
        sig_label != "",
      TRUE,
      FALSE
    )
  )

#-----------------------------
# 3. Break-even threshold
#-----------------------------
break_even_kgha <- roi_plot_df %>%
  dplyr::filter(!is.na(break_even_kgha)) %>%
  dplyr::summarise(x = dplyr::first(break_even_kgha)) %>%
  dplyr::pull(x)

#-----------------------------
# 4. Main figure
#-----------------------------
p_decision_window <- ggplot(
  roi_plot_df,
  aes(
    x = risk_class,
    y = delta_yield_kgha,
    color = product,
    group = product
  )
) +
  geom_hline(yintercept = 0, linewidth = 0.35) +
  geom_hline(yintercept = break_even_kgha, linewidth = 0.45, linetype = "dashed") +
  geom_line(linewidth = 0.6, alpha = 0.65, position = position_dodge(width = 0.35)) +
  geom_point(
    aes(shape = highlight),
    size = 3,
    stroke = 0.8,
    position = position_dodge(width = 0.35)
  ) +
  geom_text(
    data = roi_plot_df %>% dplyr::filter(highlight),
    aes(label = sig_label),
    vjust = -1.0,
    size = 4,
    show.legend = FALSE,
    position = position_dodge(width = 0.35)
  ) +
  facet_wrap(~ sds_tol, ncol = 1) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), guide = "none") +
  labs(
    title = "Decision window for SDHI seed treatment yield response",
    x = "Pre-plant pathogen-defined risk level",
    y = expression(Delta*"yield relative to CB (kg "*ha^{-1}*")"),
    color = "Seed treatment"
  ) +
  annotate(
    "text",
    x = Inf,
    y = break_even_kgha,
    label = paste0("Break-even = ", round(break_even_kgha, 0), " kg ha\u207B\u00B9"),
    hjust = 1.05,
    vjust = -0.6,
    size = 3.2
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", size = 12),
    legend.position = "top",
    legend.title = element_text(face = "bold")
  )

p_decision_window

ggsave(
  "fig_decision_window_delta_yield.png",
  p_decision_window,
  width = 9,
  height = 7,
  dpi = 600
)

#-----------------------------
# 5. Supplemental table with ROI and significance context
#-----------------------------
pairwise_sig_table <- pairs_clean %>%
  mutate(
    risk_class = as.character(risk_class),
    sds_tol = as.character(sds_tol),
    contrast = as.character(contrast),
    p_fmt = case_when(
      p.value < 0.0001 ~ "<0.0001",
      TRUE ~ sprintf("%.4f", p.value)
    )
  ) %>%
  select(
    risk_class,
    sds_tol,
    contrast,
    estimate,
    SE,
    df,
    t.ratio,
    p.value,
    p_fmt,
    sig
  )

roi_table_supp <- decision_roi_summary_fmt %>%
  mutate(
    risk_class = as.character(risk_class),
    sds_tol = as.character(sds_tol),
    product = as.character(product),
    roi_status = as.character(roi_status),
    p_value_note = case_when(
      risk_class == "3-5.Low" &
        product == "Saltro" ~ "Significant yield advantage for Saltro in pairwise contrasts",
      TRUE ~ ""
    )
  ) %>%
  arrange(
    factor(risk_class, levels = risk_order),
    sds_tol,
    factor(product, levels = c("Base", "ILEVO", "Saltro"))
  )

#-----------------------------
# 6. Export supplemental tables
#-----------------------------
output_dir <- "decision_scenario_outputs"

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

write_csv(
  roi_table_supp,
  file.path(output_dir, "supplemental_table_decision_roi_full.csv")
)

write_csv(
  pairwise_sig_table,
  file.path(output_dir, "supplemental_table_decision_pairwise_contrasts.csv")
)

#-----------------------------
# 7. Optional compact table for manuscript text
#-----------------------------
decision_window_summary <- roi_table_supp %>%
  dplyr::mutate(
    risk_class_chr = as.character(risk_class),
    product_chr = as.character(product)
  ) %>%
  dplyr::filter(
    risk_class_chr == "3-5.Low",
    stringr::str_detect(
      product_chr,
      stringr::regex("Base|CB|ILEVO|fluopyram|Saltro|pydiflumetofen", ignore_case = TRUE)
    )
  ) %>%
  dplyr::select(
    risk_class,
    sds_tol,
    product,
    emmean,                 # <-- FIXED
    delta_yield_kgha,
    net_return_usd_ha,
    roi_status
  )

readr::write_csv(
  decision_window_summary,
  file.path(output_dir, "main_text_decision_window_summary.csv")
)

decision_window_summary <- decision_window_summary %>%
  dplyr::rename(
    predicted_yield_kgha = emmean
  )

decision_window_summary


```

### freezes analysis state ###
```{r}
cache_dir <- file.path(getwd(), "cache")

if (!dir.exists(cache_dir)) {
  dir.create(cache_dir, recursive = TRUE)
}

saveRDS(m_yield_decision, file.path(cache_dir, "m_yield_decision.rds"))
saveRDS(decision_roi_summary_fmt, file.path(cache_dir, "roi_summary.rds"))
saveRDS(pairs_decision_sig, file.path(cache_dir, "pairs_sig.rds"))
```


### plotz ###
```{r}

m_yield_decision <- readRDS(file.path(cache_dir, "m_yield_decision.rds"))
decision_roi_summary_fmt <- readRDS(file.path(cache_dir, "roi_summary.rds"))
pairs_decision_sig <- readRDS(file.path(cache_dir, "pairs_sig.rds"))

#========================================================
# Main figure: predicted yield by collapsed SDS risk level
#========================================================

mean_price_usd_kg <- 0.43
break_even_kgha <- 71

#-----------------------------
# Choose predicted yield column
#-----------------------------
yield_col <- dplyr::case_when(
  "predicted_yield_kgha" %in% names(decision_roi_summary_fmt) ~ "predicted_yield_kgha",
  "pred_yield_kgha" %in% names(decision_roi_summary_fmt) ~ "pred_yield_kgha",
  "emmean" %in% names(decision_roi_summary_fmt) ~ "emmean",
  TRUE ~ NA_character_
)

#-----------------------------
# Build plotting dataframe
#-----------------------------
#-----------------------------
# Build plotting dataframe
#-----------------------------
fig_yield_df <- decision_roi_summary_fmt %>%
  mutate(
    pred_yield_kgha = .data[[yield_col]],
    risk_class = as.character(risk_class),
    product = as.character(product),
    sds_tol = as.character(sds_tol),

    fv_bin = stringr::str_extract(risk_class, "^[^.]+"),
    scn_bin = stringr::str_extract(risk_class, "(?<=\\.).+$"),

    risk_level = case_when(
      fv_bin %in% c("0", "0-1") & scn_bin %in% c("None", "Low") ~ "Low",
      fv_bin %in% c("1-3", "3-5") & scn_bin %in% c("None", "Low") ~ "Moderate",
      fv_bin == ">5" | scn_bin == "High" ~ "High",
      TRUE ~ NA_character_
    ),

    risk_level = factor(risk_level, levels = c("Low", "Moderate", "High")),

    sds_tol = case_when(
      sds_tol %in% c("r", "R") ~ "Moderately resistant cultivars (R)",
      sds_tol %in% c("s", "S") ~ "Susceptible cultivars (S)",
      TRUE ~ sds_tol
    ),

    product = case_when(
      stringr::str_detect(product, "Base") ~ "CB",
      stringr::str_detect(product, "ILEVO") ~ "CB + fluopyram",
      stringr::str_detect(product, "Saltro") ~ "CB + pydiflumetofen",
      TRUE ~ product
    ),

    product = factor(
      product,
      levels = c("CB", "CB + fluopyram", "CB + pydiflumetofen")
    )
  ) %>%
  filter(!is.na(risk_level))

#-----------------------------
# Mean ± SE
#-----------------------------
fig_yield_sum <- fig_yield_df %>%
  group_by(risk_level, sds_tol, product) %>%
  summarise(
    mean_yield = mean(pred_yield_kgha),
    se_yield = sd(pred_yield_kgha) / sqrt(n()),
    .groups = "drop"
  )

#-----------------------------
# Break-even shading (below threshold)
#-----------------------------
break_even_df <- fig_yield_sum %>%
  filter(product == "CB") %>%
  mutate(
    break_even_yield = mean_yield + break_even_kgha,
    xnum = as.numeric(risk_level),
    xmin = xnum - 0.48,
    xmax = xnum + 0.48,
    ymin = -Inf,
    ymax = break_even_yield
  )

#-----------------------------
# Brackets with ordering + spacing
#-----------------------------
product_offsets <- tibble::tribble(
  ~product, ~x_offset,
  "CB", -0.27,
  "CB + fluopyram", 0,
  "CB + pydiflumetofen", 0.27
)

sig_brackets <- pairs_decision_sig %>%
  mutate(
    risk_class = as.character(risk_class),
    sds_tol = case_when(
      as.character(sds_tol) %in% c("r", "R") ~ "Moderately resistant cultivars (R)",
      as.character(sds_tol) %in% c("s", "S") ~ "Susceptible cultivars (S)",
      TRUE ~ as.character(sds_tol)
    ),
    product_1 = case_when(
      str_detect(contrast, "Base - Saltro") ~ "CB",
      str_detect(contrast, "ILEVO - Saltro") ~ "CB + fluopyram"
    ),
    product_2 = "CB + pydiflumetofen",

    bracket_order = case_when(
      str_detect(contrast, "ILEVO - Saltro") ~ 1,
      str_detect(contrast, "Base - Saltro") ~ 2
    ),

    p_label = paste0("italic(P) == ", signif(p.value, 1))
  ) %>%
  filter(risk_class == "3-5.Low") %>%
  left_join(product_offsets, by = c("product_1" = "product")) %>%
  rename(x1_offset = x_offset) %>%
  left_join(product_offsets, by = c("product_2" = "product")) %>%
  rename(x2_offset = x_offset)

ymax_df <- fig_yield_df %>%
  group_by(sds_tol) %>%
  summarise(ymax = max(pred_yield_kgha), .groups = "drop")

sig_brackets <- sig_brackets %>%
  left_join(ymax_df, by = "sds_tol") %>%
  group_by(sds_tol) %>%
  arrange(bracket_order) %>%
  mutate(
    x_base = 2,
    x1 = x_base + x1_offset,
    x2 = x_base + x2_offset,
    y = ymax + 140 + 160 * (bracket_order - 1),
    y_tick = y - 40,
    label_y = y + 40
  ) %>%
  ungroup()

#-----------------------------
# Plot
#-----------------------------
p_yield_main <- ggplot(fig_yield_df,
  aes(x = risk_level, y = pred_yield_kgha, fill = product)
) +
  geom_rect(
    data = break_even_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey50",
    alpha = 0.1
  ) +
  geom_boxplot(
    position = position_dodge(0.8),
    width = 0.6,
    outlier.shape = NA
  ) +
  geom_jitter(
    shape = 21,
    aes(fill = product),
    color = "black",
    stroke = 0.3,
    position = position_jitterdodge(0.08, 0.8),
    alpha = 0.7,
    size = 2
  ) +
  # brackets
  geom_segment(data = sig_brackets,
    aes(x = x1, xend = x2, y = y, yend = y),
    inherit.aes = FALSE
  ) +
  geom_segment(data = sig_brackets,
    aes(x = x1, xend = x1, y = y_tick, yend = y),
    inherit.aes = FALSE
  ) +
  geom_segment(data = sig_brackets,
    aes(x = x2, xend = x2, y = y_tick, yend = y),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = sig_brackets,
    aes(x = (x1 + x2)/2, y = label_y, label = p_label),
    parse = TRUE,
    size = 3.2,
    inherit.aes = FALSE
  ) +

  facet_wrap(~ sds_tol, ncol = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Soybean sudden death syndrome (SDS) risk level",
    y = expression("Predicted soybean yield (kg "*ha^{-1}*")"),
    fill = "Seed treatment"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(face = "bold", margin = margin(r = 10)),
    axis.text = element_text(color = "black"),
    plot.margin = margin(t = 8, r = 8, b = 18, l = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_yield_main

#========================================================
# Export outputs
#========================================================

#-----------------------------
# Define output directory safely
#-----------------------------
output_dir <- file.path(getwd(), "figure_outputs")

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

#-----------------------------
# Save main figure
#-----------------------------
ggsave(
  filename = file.path(output_dir, "Fig_main_predicted_yield_collapsed_risk.png"),
  plot = p_yield_main,
  width = 7.5,
  height = 8.5,
  dpi = 600
)

# Optional vector version for journal submission
ggsave(
  filename = file.path(output_dir, "Fig_main_predicted_yield_collapsed_risk.pdf"),
  plot = p_yield_main,
  width = 7.5,
  height = 8.5
)

#-----------------------------
# Export summary data (useful for supplement / reproducibility)
#-----------------------------
write.csv(
  fig_yield_sum,
  file = file.path(output_dir, "Fig_main_summary_means_SE.csv"),
  row.names = FALSE
)

# Optional: export bracket stats
write.csv(
  sig_brackets,
  file = file.path(output_dir, "Fig_main_significance_brackets.csv"),
  row.names = FALSE
)

saveRDS(p_yield_main, file.path(output_dir, "Fig_main_plot_object.rds"))

```

### supp table? ###
```{r}

#-----------------------------
# Detect correct yield column
#-----------------------------
yield_col <- case_when(
  "predicted_yield_kgha" %in% names(decision_roi_summary_fmt) ~ "predicted_yield_kgha",
  "pred_yield_kgha" %in% names(decision_roi_summary_fmt) ~ "pred_yield_kgha",
  "emmean" %in% names(decision_roi_summary_fmt) ~ "emmean",
  TRUE ~ NA_character_
)

#-----------------------------
# Build table
#-----------------------------
decision_table <- decision_roi_summary_fmt %>%
  mutate(
    pred_yield_kgha = .data[[yield_col]],
    risk_class = as.character(risk_class),

    sds_tol = case_when(
      sds_tol %in% c("r", "R") ~ "Moderately resistant",
      sds_tol %in% c("s", "S") ~ "Susceptible",
      TRUE ~ as.character(sds_tol)
    ),

    product = case_when(
      str_detect(as.character(product), "Base") ~ "CB",
      str_detect(as.character(product), "ILEVO") ~ "CB + fluopyram",
      str_detect(as.character(product), "Saltro") ~ "CB + pydiflumetofen",
      TRUE ~ as.character(product)
    )
  ) %>%
  filter(
    risk_class %in% c("0.None", "3-5.Low", ">5.High")
  ) %>%
  select(
    risk_class,
    sds_tol,
    product,
    pred_yield_kgha,
    delta_yield_kgha,
    net_return_usd_ha,
    roi_status
  ) %>%
  mutate(
    pred_yield_kgha = round(pred_yield_kgha),
    delta_yield_kgha = round(delta_yield_kgha),
    net_return_usd_ha = round(net_return_usd_ha)
  )

emm_risk <- emmeans(
  m_yield_decision,
  ~ product | risk_class * sds_tol
)

cld_risk <- multcomp::cld(
  emm_risk,
  adjust = "tukey",
  Letters = letters
) %>%
  as.data.frame() %>%
  mutate(
    risk_class = as.character(risk_class),
    sds_tol = case_when(
      sds_tol %in% c("r", "R") ~ "Moderately resistant",
      sds_tol %in% c("s", "S") ~ "Susceptible",
      TRUE ~ as.character(sds_tol)
    ),
    product = case_when(
      stringr::str_detect(as.character(product), "Base") ~ "CB",
      stringr::str_detect(as.character(product), "ILEVO") ~ "CB + fluopyram",
      stringr::str_detect(as.character(product), "Saltro") ~ "CB + pydiflumetofen",
      TRUE ~ as.character(product)
    ),
    cld = stringr::str_trim(.group)
  ) %>%
  select(risk_class, sds_tol, product, cld)

decision_table_cld <- decision_table %>%
  left_join(cld_risk, by = c("risk_class", "sds_tol", "product"))

decision_table_cld

#-----------------------------
# Define output directory
#-----------------------------
output_dir <- file.path(getwd(), "table_outputs")

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

#-----------------------------
# Prepare table for export
#-----------------------------
decision_table_export <- decision_table_cld %>%
  mutate(
    risk_level = case_when(
      risk_class == "0.None" ~ "Low",
      risk_class == "3-5.Low" ~ "Moderate",
      risk_class == ">5.High" ~ "High",
      TRUE ~ risk_class
    ),
    treatment = product
  ) %>%
  select(
    risk_level,
    sds_tol,
    treatment,
    pred_yield_kgha,
    delta_yield_kgha,
    net_return_usd_ha,
    roi_status,
    cld
  ) %>%
  rename(
    `Risk level` = risk_level,
    `Cultivar tolerance` = sds_tol,
    `Treatment` = treatment,
    `Predicted yield (kg ha^-1)` = pred_yield_kgha,
    `Δ yield vs CB (kg ha^-1)` = delta_yield_kgha,
    `Net return (USD ha^-1)` = net_return_usd_ha,
    `ROI status` = roi_status,
    `Group` = cld
  )

#-----------------------------
# Write CSV
#-----------------------------
write.csv(
  decision_table_export,
  file = file.path(output_dir, "Supplemental_Table_3_SDHI_ROI.csv"),
  row.names = FALSE
)

# optional preview
decision_table_export


```



# -----------------------------
# Supplemental Table S1: cultivar deployment by state/province and year
# -----------------------------
```{r cultivar-summary-table, eval=FALSE}
# Purpose:
#   Generate a reviewer-friendly cultivar deployment table grouped by year and
#   state/province, with moderately resistant and susceptible cultivars shown
#   side-by-side. This table supports the Methods and Supplemental Table S1.
#
# Output:
#   tables/Supplemental_Table_Cultivars_by_State_Year.csv

state_lookup <- tibble::tribble(
  ~state, ~state_full,
  "DE", "Delaware",
  "IA", "Iowa",
  "IL", "Illinois",
  "IN", "Indiana",
  "KS", "Kansas",
  "KY", "Kentucky",
  "MI", "Michigan",
  "MO", "Missouri",
  "ND", "North Dakota",
  "OH", "Ohio",
  "ON", "Ontario",
  "WI", "Wisconsin"
)

cultivar_table_wide <- df %>%
  select(state, year, location, cultivar, sds_tol) %>%
  distinct() %>%
  mutate(
    state = trimws(as.character(state)),
    year = as.integer(as.character(year)),
    location = trimws(as.character(location)),
    cultivar = stringr::str_to_upper(trimws(as.character(cultivar))),
    tol_class = case_when(
      sds_tol %in% c("r", "R") ~ "moderately resistant (R) cultivar",
      sds_tol %in% c("s", "S") ~ "susceptible (S) cultivar",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(tol_class), !is.na(cultivar), cultivar != "") %>%
  group_by(year, state, tol_class) %>%
  summarise(
    cultivars = str_c(sort(unique(cultivar)), collapse = "; "),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(
    names_from = tol_class,
    values_from = cultivars,
    values_fill = "-"
  )

cultivar_counts <- df %>%
  select(state, year, location, cultivar, sds_tol) %>%
  distinct() %>%
  mutate(
    state = trimws(as.character(state)),
    year = as.integer(as.character(year)),
    location = trimws(as.character(location)),
    cultivar = stringr::str_to_upper(trimws(as.character(cultivar))),
    sds_tol = trimws(as.character(sds_tol))
  ) %>%
  filter(
    sds_tol %in% c("r", "R", "s", "S"),
    !is.na(cultivar),
    cultivar != ""
  ) %>%
  group_by(year, state) %>%
  summarise(
    n_cultivars = n_distinct(cultivar),
    n_siteyears = n_distinct(paste(location, year)),
    .groups = "drop"
  )

cultivar_table_export <- cultivar_table_wide %>%
  left_join(cultivar_counts, by = c("year", "state")) %>%
  left_join(state_lookup, by = "state") %>%
  mutate(`state/province` = if_else(is.na(state_full), state, state_full)) %>%
  select(
    year,
    `state/province`,
    `moderately resistant (R) cultivar`,
    `susceptible (S) cultivar`,
    n_cultivars,
    n_siteyears
  ) %>%
  arrange(year, `state/province`)

readr::write_csv(
  cultivar_table_export,
  file.path(table_dir, "Supplemental_Table_Cultivars_by_State_Year.csv")
)

cultivar_table_export
```


# -----------------------------
# Reproducibility record
# -----------------------------
```{r session-info}
# Record package versions and R version used to render the pipeline.
sessionInfo()
```
